---
title: "`r params$id` Clustering in separate samples"
author: "aslı"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  id: 'id'
  output_path: 'output_path' # where we save the sce with clustering data 
  path_to_sce_norm: 'sce_norm'
  chosen_perp: 'chosen_perp'
  perplexity_list: 'perplexity_list'
  k_list: 'k_list'
  chosen_k: 'chosen_k'
  n_PCs: 'n_PCs'
---
  
In this report, we perform clustering in isolated samples and show the results in dimensionality  
reduction plots. We also perform modularity analysis and bootstrap to test the quality of our clusters.  
More info about these two methods are given below, at the top of the respective plots.  
We use **scran** package for clustering. The algorithm we chose is **nearest neighbors** since it is  
widely adopted.  


```{r include = FALSE}
# source some files 
library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
```

```{r echo = FALSE, message = FALSE}

# load the data - it has to be normalized sce with reduced dimensions 
sce <- readRDS(params$path_to_sce_norm)

perplexity_list <- params$perplexity_list
perplexity_list <- strsplit(perplexity_list, " ") # perplexity_list is passed a string with space, get rid of spaces
perplexity_list <- as.list(c(unlist(list(perplexity_list)))) # save each element separately 
perplexity_list <- lapply(perplexity_list, function(element) as.numeric(element)) # make them numeric 

k_list <- params$k_list
k_list <- strsplit(k_list, " ") # perplexity_list is passed a string with space, get rid of spaces
k_list <- as.list(c(unlist(list(k_list)))) # save each element separately 
k_list <- lapply(k_list, function(element) as.numeric(element)) # make them numeric 

chosen_perp <- params$chosen_perp
chosen_perp <- as.numeric(chosen_perp)

chosen_k <- params$chosen_k
chosen_k <- as.numeric(chosen_k)

```

```{r include = FALSE}

g <- buildSNNGraph(sce, use.dimred = "PCA", k = 15) # shared neighbor weighting
cluster <- igraph::cluster_walktrap(g)$membership # extract clustering info from the graph that we just made 
sce$cluster <- factor(cluster) # add that info to the sce 
cluster_table <- table(sce$cluster) # this table shows how many cells in each cluster 

# we can redo the reduced dim plots with the clustering info in mind 
plotTSNE <- plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', params$id, sep = '_'))

```

### t-SNE plots with various perplexities  
Perplexity is an important parameter that determines the granularity of the visualization. Low perplexities will favor resolution of finer structure where local variations dominate, possibly to the point that the visualization is compromised by random noise. Larger perplexity values favor the global geometry of the dataset. Thus, it is advisable to test different perplexity values to ensure that the choice of perplexity does not drive the interpretation of the plot. The performance of SNE is fairly robust to changes in the perplexity, and typical values are between 5 and 50. Below t-SNE plots with various perplexity values are shown. t-SNE with a perplexity of `r params$chosen_perp` is chosen for further downstream analysis.  
  
Note that in the plots below, shared neighbor algorithm was used with 15 neighbors. 

```{r, message = FALSE, echo = FALSE}

# perplexity_list <- list(5, 10, 15, 20, 30, 40, 50, 60, 70, 80, 90, 100)
# we will use this to dynamically change the figure height in the next chunk
dim <- length(perplexity_list) / 3 * 4

```

```{r fig.width = 12, fig.height = dim, message = FALSE}

# define a function to make tsne plots 
make_tsne_plot <- function(sce, perplexity_value) {
  
  sce <- runTSNE(sce, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, perplexity = perplexity_value)
  p <- list(plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste("perplexity = ", perplexity_value, sep = "")))
  
  return(p)
}

# make copies of the sce object, as many times as the number of different perplexities supplied 
sce_list <- list()
i = 1 
while (i <= length(perplexity_list)) {
  
  new_sce <- sce # make a copy of the sce 
  sce_list[[i]] <- new_sce
  
  i = i + 1 
}

# list of tsne plots with various perplexity values 
plotlist <- mapply(make_tsne_plot, sce = sce_list, perplexity_value = perplexity_list)

# based on the user passed chosen_perp parameter, pick an sce computed with a certain perplexity value 
index <- match(chosen_perp, perplexity_list) # find the index of the chosen perp in the perp list 
sce <- sce_list[[index]]

# visualize all the plots in 3 columns
do.call(gridExtra::grid.arrange, c(plotlist, ncol = 3))

```

### t-SNE plots with various numbers of clusters 
After picking a t-SNE plot with a certain perplexity (`r params$chosen_perp`), next we try various numbers for nearest neighbor to see how this changes the number of clusters we have. 

```{r, include = FALSE}

# k_list = list(5, 9, 10, 12, 15, 18)

# we will use this to dynamically change the figure height in the next chunk
dim <- length(k_list) / 3 * 4

```


```{r fig.width = 12, fig.height = dim}

# make copies of the sce object, as many times as the number of different perplexities supplied 
sce_list <- list()
i = 1 
while (i <= length(k_list)) {
  
  new_sce <- sce # make a copy of the sce 
  sce_list[[i]] <- new_sce
  
  i = i + 1 
}

# function to cluster sce with various k values
cluster_sce <- function(sce, k_value) {
  
  g <- buildSNNGraph(sce, use.dimred = "PCA", k = k_value, type = "number") # shared neighbor weighting
  cluster <- igraph::cluster_louvain(g)$membership # extract clustering info from the graph that we just made 
  sce$cluster <- factor(cluster) # add that info to the sce 

  return(sce)
  
}

# function to make tsne plots for various sces 
make_tsne_plots <- function(sce, k_value) {
  
  p <- list(plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste("k = ", k_value, sep = "")))

  return(p)
}

sce_list_cluster <- mapply(cluster_sce, sce = sce_list, k_value = k_list) # generate list of sces with various numbers of clusters 
tsne_plots <- mapply(make_tsne_plots, sce = sce_list_cluster, k_value = k_list) # generate tsne plots using the sce_list and various k values

# based on the user passed chosen_k parameter, pick an sce computed with a certain k value 
index <- match(chosen_k, k_list) # find the index of the chosen k in the k list 
sce <- sce_list_cluster[[index]]

do.call(gridExtra::grid.arrange, c(tsne_plots, ncol = 3))

```
### PCA  
By inspecting the contribution of each PC to overall variance, we can get a better sense of the heterogeneity of the data.  
```{r fig.width = 12, fig.height = 4}

percent.var <- attr(reducedDim(sce), "percentVar")

# visualize first 100 PCs
if (is.null(percent.var)){
  
  message(print("Can't produce the plot for this sample. Refer to other methods below to decide how many PCs to include."))
  
} else {plot(percent.var[1:100], xlab = "PC", ylab = "Variance explained (%)", type = "p", pch = 16, main = "contribution of top 100 PCs to variance") 

}

```
We can also plot the top 4 PCs together as shown below to visualize multiple PCs at the same time. Cells grouped according to the cluster they belong to. The diagonal boxes in the scatter plot matrix below show the cell density for each component.  

```{r fig.width = 12, fig.height = 8}

plotReducedDim(sce, dimred="PCA", ncomponents = 4, colour_by="cluster") + ggtitle("4 top PCs")

```

Final dimensionality reduction plots shown below with clustering information: 
```{r fig.width = 12, fig.height = 8}

p1 <- plotPCA(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', params$id, sep = '_'))
p2 <- plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('t-SNE', params$id, sep = '_'))
p3 <- plotUMAP(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', params$id, sep = '_'))

gridExtra::grid.arrange(p1, p2, p3, ncol = 2)

```

### Diffusion maps  
Diffusion maps are a non-linear dimensionality technique. It achieves dimensionality reduction by re-organising data according to parameters of its underlying geometry. This method can help us a lot when dealing with data that includes some sort of differentiation or transformation process because diffusion maps are really good at picking up brancing in the data. This is a computationally expensive calculation, so to speed things up, it's best to limit the computation to most significant PCs. In this report, top `r params$n_PCs` were used. 

```{r fig.width = 12, fig.height = 8, message = FALSE}

message(print("Computing the diffusion map"))

topPCs <- as.numeric(params$n_PCs)

sce_logcounts <- logcounts(sce)  # access log-transformed counts matrix
colnames(sce_logcounts) <- sce$cluster
sce_logcounts <- as.matrix(sce_logcounts)

# Make a diffusion map, this step can take a while depending on how large the sce object is, up to 10 minutes. 
dm <- DiffusionMap(t(sce_logcounts), n_pcs = 30)

tmp <- data.frame(DC1 = eigenvectors(dm)[, 1],
                  DC2 = eigenvectors(dm)[, 2],
                  Timepoint = sce$cluster)

dm_plot <- ggplot(tmp, aes(x = DC1, y = DC2, colour = Timepoint)) +
  geom_point() + 
  xlab("Diffusion component 1") + 
  ylab("Diffusion component 2") +
  theme_classic() + 
  ggtitle("Diffusion map")

sce$pseudotime_diffusionmap <- rank(eigenvectors(dm)[,1])    # rank cells by their dpt

ggplot(as.data.frame(colData(sce)), 
       aes(x = pseudotime_diffusionmap, 
           y = cluster, colour = cluster)) +
    geom_quasirandom(groupOnX = FALSE) +
    theme_classic() +
    xlab("Diffusion component 1 (DC1)") + ylab("Timepoint") +
    ggtitle("Cells ordered by DC1")

plot(eigenvalues(dm), ylim = 0:1, pch = 20, xlab = 'Diffusion component (DC)', ylab = 'Eigenvalue')


plot(dim, pch = 20, col.by = "cluster")

sigmas <- find_sigmas(sce_logcounts, verbose = FALSE)
optimal.sigma(sigmas)


```





### Dimensionality reduction using Seurat  
Plots shown above were generated using Scater and Scran R packages. We can also check out Seurat and compare the results across various packages. The main challenge is to decide how many PCs to include in further analysis. Seurat gives a couple of tools that we will explore further. These tools are:  
 - Heatmaps  
 - Making use of p values   
 - Finding the elbow point  

Below the distribution of cells in top 6 PCs is visualized where cells are colored according to the cluster they belong to.

```{r include = FALSE}

# convert sce to seurat object
sobject <- as.Seurat(sce, counts = "counts", data = "logcounts")
Idents(sobject) <- paste(unique(sce$id))

sobject <- ScaleData(sobject)
sobject <- FindVariableFeatures(sobject)

sobject@reductions$pca <- sobject@reductions$PCA
sobject@reductions$tsne <- sobject@reductions$TSNE
sobject@reductions$umap <- sobject@reductions$UMAP

sobject <- RunPCA(sobject, features = VariableFeatures(object = sobject))

```


```{r echo = FALSE, message = FALSE, fig.width = 12, fig.height = 4}

# Visualize top 6 PCs 
gridExtra::grid.arrange(DimPlot(sobject, reduction = "pca", group.by = "cluster",dims = 1:2),
                        DimPlot(sobject, reduction = "pca", group.by = "cluster",dims = 3:4),
                        DimPlot(sobject, reduction = "pca", group.by = "cluster",dims = 5:6), ncol = 3, top = 'top 6 PCs visualized')

```

#### Visualize top genes associated with reduction components, in top 6 PCs
```{r, fig.width = 14, fig.height = 10}

VizDimLoadings(sobject, dims = 1:6, reduction = "pca", ncol = 3)

```

#### Deciding how many PCs to include in further analysis  
In the following heatmaps top genes contributing to each PC are visualized. Both cells and genes are ordered by their PC scores, and the 15 genes with the highest and 15 genes with the lowest PC scores are displayed for each PC. Note that the cells are given in the columns and genes are show in the rows. Here the goal is to pick the PCs that can divide the cell population into distinct groups. If all of the cells are the same color (pink), then that PC isn't as significant. 

```{r fig.width = 16, fig.height = 25}

DimHeatmap(sobject, dims = 1:15, cells = 500, balanced = TRUE)

```

#### Finding significant PCs
‘Significant’ PCs will show a strong enrichment of genes with low p-values (solid curve above the dashed line) in the plot below.  

```{r fig.width = 10, fig.height = 7, warning = FALSE, message=FALSE}

sobject <- JackStraw(object = sobject, reduction = "pca", dims = 20, num.replicate = 100,  prop.freq = 0.1, verbose = FALSE)

sobject <- ScoreJackStraw(object = sobject, dims = 1:20, reduction = "pca")
JackStrawPlot(object = sobject, dims = 1:20, reduction = "pca")


```

To further validate the results of the plot above, we can also check out contribution of each PC to total standard deviation and find the elbow point.  
```{r fig.width = 10, fig.height = 4}

ElbowPlot(object = sobject)

```


```{r include = FALSE}

# here we save the sce object with the clustering data - we only save the shared neighbor clusters 
saveRDS(sce, file = params$output_path)

```

```{r echo= FALSE, message = FALSE}
# now we make the table look nice for the report 

cluster_table <- as.data.frame(cluster_table) %>% 
  rename(cluster = Var1, number_of_cells = Freq)

cluster_table %>% 
  knitr::kable(caption = 'Table 1: number of cells in each cluster - shared neighbor') %>% 
  kable_styling(full_width = F)

```

```{r include = FALSE}
# CHECK THE ACCURACY OF OUR CLUSTERING: 
# here we calculate the modularity scores for our clusters. this helps understand the relationships between clusters. 
# we want the clusters to be distinct, so we dont want different clusters to interact closely. 
ratio <- clusterModularity(g, cluster, as.ratio=TRUE)

# here, we want the diagonal squares to be dark but the other ones to be as light as possible. 
# darker colors mean higher modularity - higher interaction 
# this will be helpful to diagnose low quality clustering 
heatMap_modularity <- pheatmap(log10(ratio+1), cluster_cols=FALSE, cluster_rows=FALSE, col=rev(heat.colors(100)), main = 'modularity')

# now we do bootstrapping on our clusters. ideally, a cell in a cluster should always remain there. 
boot_results <- bootstrapCluster(sce, FUN=function(x) {
  g <- buildSNNGraph(x, use.dimred="UMAP")
  igraph::cluster_walktrap(g)$membership}, clusters=sce$cluster)

# same thing, you only want the diagonal squares to have a dark color. 
heatMap_bootstrap <- pheatmap(boot_results, cluster_cols=FALSE, cluster_rows=FALSE, col=colorRampPalette(c("white", "blue"))(100), main = 'bootstrap')
```


#### Modularity scores 
Here, we measure how much our clusters interact. Ideally, we want the interaction within a cluster  
to be high, but we don't want separate clusters to interact. This heatmap helps visualize this:  
the diagonal squares should be darkest, but the surrounding squares should be as light as possible.  
If you see a lot of darker squares around, you may want to reconsider your clustering parameters.  
The lower part of the heatmap is obviously not applicable, just focus on the top part of the plot.   
Note that the darker the square, stronger the interaction.

```{r echo = FALSE, message = FALSE}
heatMap_modularity
```


#### Bootstrap results 
We use bootstrapping multiple times to redo the clustering. Ideally, cells in a cluster would remain there  
in each time we do the clustering again. The way you read this heatmap is very similar to the heatmap above.  
The best case is when the diagonal squares are darkest but the other ones are as light as possible.  
If you see many squares with darker colors outside the main diagonal, that means cells appeared in  
various clusters as we redid clustering through bootstrapping. 

```{r echo = FALSE, message = FALSE}
heatMap_bootstrap
```

### ALGORITHM: k means 
```{r include = FALSE}
# Build a k means graph for cells based on their expression profiles.
g <- buildKNNGraph(sce, use.dimred="UMAP") # k means 

# extract clustering info from the graph that we just made 
cluster <- igraph::cluster_walktrap(g)$membership

# add that info to the sce 
sce$cluster <- factor(cluster)

# this table shows how many cells in each cluster 
cluster_table <- table(sce$cluster)

# we can redo the reduced dim plots with the clustering info in mind 
plotTSNE <- plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', params$id, sep = '_'))
plotUMAP <- plotUMAP(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', params$id, sep = '_'))
plotPCA <- plotPCA(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', params$id, sep = '_'))

```

```{r echo= FALSE, message = FALSE}
# now we make the table look nice for the report 

cluster_table <- as.data.frame(cluster_table) %>% 
  rename(cluster = Var1, number_of_cells = Freq)

cluster_table %>% 
  knitr::kable(caption = 'Table 1: number of cells in each cluster') %>% 
  kable_styling(full_width = F)

```

```{r echo = FALSE, message = FALSE}
# visualize these graphs 
plotTSNE
plotUMAP
plotPCA
```


```{r include = FALSE}
# CHECK THE ACCURACY OF OUR CLUSTERING: 
# here we calculate the modularity scores for our clusters. this helps understand the relationships between clusters. 
# we want the clusters to be distinct, so we dont want different clusters to interact closely. 
ratio <- clusterModularity(g, cluster, as.ratio=TRUE)

# here, we want the diagonal squares to be dark but the other ones to be as light as possible. 
# darker colors mean higher modularity - higher interaction 
# this will be helpful to diagnose low quality clustering 
heatMap_modularity <- pheatmap(log10(ratio+1), cluster_cols=FALSE, cluster_rows=FALSE, col=rev(heat.colors(100)), main = 'modularity')

# now we do bootstrapping on our clusters. ideally, a cell in a cluster should always remain there. 
boot_results <- bootstrapCluster(sce, FUN=function(x) {
  g <- buildSNNGraph(x, use.dimred="UMAP")
  igraph::cluster_walktrap(g)$membership}, clusters=sce$cluster)

# same thing, you only want the diagonal squares to have a dark color. 
heatMap_bootstrap <- pheatmap(boot_results, cluster_cols=FALSE, cluster_rows=FALSE, col=colorRampPalette(c("white", "blue"))(100), main = 'bootstrap')
```

#### Modularity scores 

```{r echo = FALSE, message = FALSE}
heatMap_modularity
```

#### Bootstrap results 

```{r echo = FALSE, message = FALSE}
heatMap_bootstrap
```







