---
title: "`r params$id` Clustering in separate samples"
author: "aslı"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  id: 'DH16'
  output_path: '/huntsman/amunzur/data/clustered/sce/DH16/sce_clus.rds' # where we save the sce with clustering data 
---
  
In this report, we perform clustering in isolated samples and show the results in dimensionality  
reduction plots. We also perform modularity analysis and bootstrap to test the quality of our clusters.  
More info about these two methods are given below, at the top of the respective plots.  
We use **scran** package for clustering. The algorithm we chose is **nearest neighbors** since it is  
widely adopted. Scran picks the best number of clusters and uses graph based clustering methods.  


```{r include = FALSE}
# source some files 
library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
```

```{r include = FALSE}
# load the data - it has to be normalized sce with reduced dimensions 
sce <- readRDS(here('..', 'data', 'normalized', params$id, 'sce_norm.rds'))
```

### ALGORITHM: Shared neighbor
```{r include = FALSE}
# Build a shared neighbor graph for cells based on their expression profiles.
g <- buildSNNGraph(sce, use.dimred="UMAP") # shared neighbor weighting

# extract clustering info from the graph that we just made 
cluster <- igraph::cluster_walktrap(g)$membership

# add that info to the sce 
sce$cluster <- factor(cluster)

# this table shows how many cells in each cluster 
cluster_table <- table(sce$cluster)

# we can redo the reduced dim plots with the clustering info in mind 
plotTSNE <- plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', params$id, sep = '_'))
plotUMAP <- plotUMAP(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', params$id, sep = '_'))
plotPCA <- plotPCA(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', params$id, sep = '_'))

```

```{r include = FALSE}
# here we save the sce object with the clustering data - we only save the shared neighbor clusters 
saveRDS(sce, file = params$output_path)

```


```{r echo= FALSE, message = FALSE}
# now we make the table look nice for the report 

cluster_table <- as.data.frame(cluster_table) %>% 
  rename(cluster = Var1, number_of_cells = Freq)

cluster_table %>% 
  knitr::kable(caption = 'Table 1: number of cells in each cluster - shared neighbor ') %>% 
  kable_styling(full_width = F)

```

```{r echo = FALSE, message = FALSE}
# visualize these graphs 
plotTSNE
plotUMAP
plotPCA
```

```{r include = FALSE}
# CHECK THE ACCURACY OF OUR CLUSTERING: 
# here we calculate the modularity scores for our clusters. this helps understand the relationships between clusters. 
# we want the clusters to be distinct, so we dont want different clusters to interact closely. 
ratio <- clusterModularity(g, cluster, as.ratio=TRUE)

# here, we want the diagonal squares to be dark but the other ones to be as light as possible. 
# darker colors mean higher modularity - higher interaction 
# this will be helpful to diagnose low quality clustering 
heatMap_modularity <- pheatmap(log10(ratio+1), cluster_cols=FALSE, cluster_rows=FALSE, col=rev(heat.colors(100)), main = 'modularity')

# now we do bootstrapping on our clusters. ideally, a cell in a cluster should always remain there. 
boot_results <- bootstrapCluster(sce, FUN=function(x) {
  g <- buildSNNGraph(x, use.dimred="UMAP")
  igraph::cluster_walktrap(g)$membership}, clusters=sce$cluster)

# same thing, you only want the diagonal squares to have a dark color. 
heatMap_bootstrap <- pheatmap(boot_results, cluster_cols=FALSE, cluster_rows=FALSE, col=colorRampPalette(c("white", "blue"))(100), main = 'bootstrap')
```


#### Modularity scores 
Here, we measure how much our clusters interact. Ideally, we want the interaction within a cluster  
to be high, but we don't want separate clusters to interact. This heatmap helps visualize this:  
the diagonal squares should be darkest, but the surrounding squares should be as light as possible.  
If you see a lot of darker squares around, you may want to reconsider your clustering parameters.  
The lower part of the heatmap is obviously not applicable, just focus on the top part of the plot.   
Note that the darker the square, stronger the interaction.

```{r echo = FALSE, message = FALSE}
heatMap_modularity
```


#### Bootstrap results 
We use bootstrapping multiple times to redo the clustering. Ideally, cells in a cluster would remain there  
in each time we do the clustering again. The way you read this heatmap is very similar to the heatmap above.  
The best case is when the diagonal squares are darkest but the other ones are as light as possible.  
If you see many squares with darker colors outside the main diagonal, that means cells appeared in  
various clusters as we redid clustering through bootstrapping. 

```{r echo = FALSE, message = FALSE}
heatMap_bootstrap
```

### ALGORITHM: k means 
```{r include = FALSE}
# Build a k means graph for cells based on their expression profiles.
g <- buildKNNGraph(sce, use.dimred="UMAP") # k means 

# extract clustering info from the graph that we just made 
cluster <- igraph::cluster_walktrap(g)$membership

# add that info to the sce 
sce$cluster <- factor(cluster)

# this table shows how many cells in each cluster 
cluster_table <- table(sce$cluster)

# we can redo the reduced dim plots with the clustering info in mind 
plotTSNE <- plotTSNE(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', params$id, sep = '_'))
plotUMAP <- plotUMAP(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', params$id, sep = '_'))
plotPCA <- plotPCA(sce, colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', params$id, sep = '_'))

```




```{r echo= FALSE, message = FALSE}
# now we make the table look nice for the report 

cluster_table <- as.data.frame(cluster_table) %>% 
  rename(cluster = Var1, number_of_cells = Freq)

cluster_table %>% 
  knitr::kable(caption = 'Table 1: number of cells in each cluster') %>% 
  kable_styling(full_width = F)

```

```{r echo = FALSE, message = FALSE}
# visualize these graphs 
plotTSNE
plotUMAP
plotPCA
```


```{r include = FALSE}
# CHECK THE ACCURACY OF OUR CLUSTERING: 
# here we calculate the modularity scores for our clusters. this helps understand the relationships between clusters. 
# we want the clusters to be distinct, so we dont want different clusters to interact closely. 
ratio <- clusterModularity(g, cluster, as.ratio=TRUE)

# here, we want the diagonal squares to be dark but the other ones to be as light as possible. 
# darker colors mean higher modularity - higher interaction 
# this will be helpful to diagnose low quality clustering 
heatMap_modularity <- pheatmap(log10(ratio+1), cluster_cols=FALSE, cluster_rows=FALSE, col=rev(heat.colors(100)), main = 'modularity')

# now we do bootstrapping on our clusters. ideally, a cell in a cluster should always remain there. 
boot_results <- bootstrapCluster(sce, FUN=function(x) {
  g <- buildSNNGraph(x, use.dimred="UMAP")
  igraph::cluster_walktrap(g)$membership}, clusters=sce$cluster)

# same thing, you only want the diagonal squares to have a dark color. 
heatMap_bootstrap <- pheatmap(boot_results, cluster_cols=FALSE, cluster_rows=FALSE, col=colorRampPalette(c("white", "blue"))(100), main = 'bootstrap')
```


#### Modularity scores 

```{r echo = FALSE, message = FALSE}
heatMap_modularity
```


#### Bootstrap results 

```{r echo = FALSE, message = FALSE}
heatMap_bootstrap
```







