---
title: "`r params$ids` Clustering analysis on combined samples"
author: "asli"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  ids: 'pair_ids'
  output_uncorrected: 'output_uncorrected'
  output_integrated: 'output_integrated'
---
This report shows the results of clustering after we combine the samples.  
We first show the results of clustering within each sample, then we compare those  
results to clustering in a combined sample. For clustering in a combined sample,  
we first do clustering analysis on individual samples by finding neighbors.  
Based on our findings, we then define clusters. Seurat package picks the best number  
of clusters for us. The algorithm we use is **nearest neighbors**.  
When we have this data for both samples, we follow these steps:  

1. find integration anchors, these are cells that have similar biological state  
across samples  
2. integrate the the samples into one larger sample  
3. run dimension reduction analyses on the larger sample as well as a new clustering analysis  
(best number of clusters is again decided by seurat)

Note that this script is also good for just integrating two samples and removing batch effects. 


```{r include = FALSE}
# source the files we need. this file also contains the packages we need. 
library(here)
source(here('..', 'sourceFiles', 'utilities.R'))

```


```{r include = FALSE}
# extract our data from the ids 
ids <- strsplit(params$ids, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
dh_organoid_ids <- list(ids[[1]], ids[[2]])
```

```{r include = FALSE}
# load the sce objects we will be working on 
sce_list <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'clustered', 'sce', id, 'sce_clus.rds')))

#name them nicely 
names(sce_list) <- dh_organoid_ids
```

### We fist visualize the reduced dim plots (tsne and umap) separately for each sample.   
#### First data set 

```{r include = FALSE}

a <- plotTSNE(sce_list[[1]], colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', dh_organoid_ids[[1]], sep = '_'))
b <- plotUMAP(sce_list[[1]], colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', dh_organoid_ids[[1]], sep = '_'))

# plotPCA(sce_list[[1]], colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', dh_organoid_ids[[1]], sep = '_'))
```

#### Second data set  
```{r include = FALSE}

c <- plotTSNE(sce_list[[2]], colour_by="cluster", text_by="cluster") + ggtitle(paste('TSNE', dh_organoid_ids[[2]], sep = '_'))
d <- plotUMAP(sce_list[[2]], colour_by="cluster", text_by="cluster") + ggtitle(paste('UMAP', dh_organoid_ids[[2]], sep = '_'))
# plotPCA(sce_list[[2]], colour_by="cluster", text_by="cluster") + ggtitle(paste('PCA', dh_organoid_ids[[2]], sep = '_'))
```

```{r echo = FALSE, message = FALSE}

gridExtra::grid.arrange(a, b, c, d, nrow = 2)

```

Frequency table: 
This table shows the frequency of cell id types in each sample.
```{r echo = FALSE, message = FALSE}

# find the common genes between the sces 
universal <- Reduce(intersect, list(rownames(sce_list[[1]]), rownames(sce_list[[2]])))

# only keep the common genes in all sces 
sce_list <- lapply(sce_list, function(sce) sce <- sce[universal, ])

# we run this function on our sces here. we just convert them into seurat objects and do a bit of a clean up. function is defined in the utilities file. 
sobject1 <- seuratClustering(sce_list[[1]], paste(dh_organoid_ids[[1]]))
sobject2 <- seuratClustering(sce_list[[2]], paste(dh_organoid_ids[[2]]))

# LETS START OUR INTEGRATION:
# first combine the objects and add cell ids as well so that we know where each cell came from 
DH.combined = merge(sobject1, y = sobject2, add.cell.ids = c(dh_organoid_ids[[1]], dh_organoid_ids[[2]]), project = "protocol")

# save this object, snakemake passes the filepath for the output object 
saveRDS(DH.combined, file = 'output_uncorrected')

# lets split by protocol 
data.list <- SplitObject(DH.combined, split.by = "protocol")

# make a reference list for our samples, we'll use it in the next step 
reference.list <- data.list[c(paste(dh_organoid_ids[[1]]), dh_organoid_ids[[2]])]

# this is where things get messy! we will follow a different path depending on how many cells we have in each sample. 
# FindIntegrationAnchors will be used only if we have at least 200 cells in each sample 

# condition based on cell number: do this first method if we have more than 200 cells 
if (dim(DH.combined)[[2]] > 200) {
  
  # find anchoring points between two samples and put them in the refence list, use first 30 dimensions. 
  # these anchors represent cells with similar states between two samples 
  # seurat uses those samples to normalize the data. 
  # when we have less than 200 cells, we will use other normalization methods for batch normalization 
  samples.anchors <- FindIntegrationAnchors(object.list = reference.list, dims = 1:30, anchor.features = 3000 )
  
  # integrate the two data sets 
  samples.integrated <- IntegrateData(anchorset = samples.anchors, dims = 1:30, features.to.integrate = rownames(DH.combined))
  
  # set the default combined assay we will be using. we will be using the 'integrated', aka batch corrected assay, not the previous ones 
  DefaultAssay(samples.integrated) <- "integrated"
  
  # do some gene level scaling. we scale so that each gene will be centered to have a mean of 0 and and scaled by SD of the gene
  samples.integrated <- ScaleData(samples.integrated, verbose = FALSE)
  
  # run some dim reduction 
  samples.integrated <- RunPCA(samples.integrated, npcs = 30, verbose = FALSE)
  samples.integrated <- RunTSNE(samples.integrated, dims = 1:30)
  samples.integrated <- RunUMAP(samples.integrated, dims = 1:30)
  
  # make the plots here 
  # this is how our data basically looks like when we combine: 
  # the group.by command shows the differences between groups based on protocol
  p1 <- DimPlot(samples.integrated, reduction = "tsne", group.by = "protocol")
   
  # this command shows the same integrated object, but it separates it based on clusters 
  p2 <- DimPlot(samples.integrated, reduction = "tsne", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('TSNE')
  
  p3 <- DimPlot(samples.integrated, reduction = "umap", group.by = "protocol")
  p4 <- DimPlot(samples.integrated, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('UMAP')
  
  p5 <- DimPlot(samples.integrated, reduction = "pca", group.by = "protocol")
  p6 <- DimPlot(samples.integrated, reduction = "pca", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('PCA')
  
  # this piece of code will show you the frequency of cell id types in each sample.
  freq_table <- prop.table(x = table(samples.integrated@active.ident, samples.integrated@meta.data[, "protocol"]), margin = 2)
  
  freq_table %>% 
    knitr::kable(caption = 'Table 1: frequency of each cell type (cluster) in each sample') %>% 
    kable_styling(full_width = F)

  
} else { # this is what to do if yo uhave less than 200 samples: 
  
  # subset to the common genes: 
  universal <- intersect(rownames(sce_list[[1]]), rownames( sce_list[[2]]))
  sce_list[[1]] <- sce1[universal,]
  sce_list[[2]] <- sce2[universal,]
  
  # remove differences in sequencing depth between batches  
  rescaled <- multiBatchNorm(sce_list[[1]], sce_list[[2]]) 
  
  # update our sces 
  sce_list[[1]] <- rescaled[[1]]
  sce_list[[2]] <- rescaled[[2]]
  
  # integration using nearest neighbors
  corrected <- fastMNN(one = sce_list[[1]], two =  sce_list[[2]], subset.row=chosen.hvgs)
  
  # combine counts and log counts to make a combined sce 
  overall_counts <- cbind(counts(sce_list[[1]]), counts( sce_list[[2]]))
  overall_logcounts <- cbind(logcounts(sce_list[[1]]), logcounts( sce_list[[2]]))
  
  # do the combining here
  sce <- SingleCellExperiment( 
    
    assays = list(counts = overall_counts, logcounts = overall_logcounts),  
    rowData = rowData(sce_list[[1]]), # same as rowData(pbmc4k) 
    colData = rbind(colData(sce_list[[1]]), colData( sce_list[[2]])))

  # run reduced dim analysis 
  # set the seed for reproducibility
  set.seed(seed)
  
  sce <- runPCA(sce, exprs_values = "logcounts", ncomponents = 200)
  sce <- runTSNE(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3)
  sce <- runUMAP(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3,
                    min_dist = 0.01, n_neighbors = 15, metric = "euclidean")

  # convert to seurat object 
  combined_seurat <- as.seurat(sce)
  
  # build a snn graph 
  combined_seurat <-FindNeighbors(combined_seurat)
  
  # lets find some clusters 
  combined_seurat <- FindClusters(combined_seurat, resolution = 0.15)
  
  # run reduced dim analysis / i probably dont need this? 
  # sce <- RunPCA(sce, npcs = 30, verbose = FALSE)
  # sce <- RunTSNE(sce, dims = 1:30)
  # sce <- RunUMAP(sce, dims = 1:30)
  
  # do some gene level scaling. we scale so that each gene will be centered to have a mean of 0 and and scaled by SD of the gene
  samples.integrated <- ScaleData(samples.integrated, verbose = FALSE)
  
  # make some reduced dim plots 
  
  # the group.by command shows the differences between groups based on id
  p1 <- DimPlot(samples.integrated, reduction = "tsne", group.by = "id")
   
  # this command shows the same integrated object, but it separates it based on clusters 
  p2 <- DimPlot(samples.integrated, reduction = "tsne", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('TSNE')
  
  p3 <- DimPlot(samples.integrated, reduction = "umap", group.by = "id")
  p4 <- DimPlot(samples.integrated, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('UMAP')
  
  # p5 <- DimPlot(samples.integrated, reduction = "pca", group.by = "id")
  # p6 <- DimPlot(samples.integrated, reduction = "pca", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend() + ggtitle('PCA')
  
  ##################################################################################################################
  ##################################################################################################################
  
} # end of else 

# save the integrated object, snakemake passes the filepath for saving 
saveRDS(samples.integrated, file = 'output_integrated')

```


The following plots are generated after we integrate the two samples and run a clustering algorithm on the integrated sample. 
Now, we run a clustering analysis on the combined data and visualize our findings in dimension reduction plots.  
Each cluster is shown with a different color.  

#### TSNE PLOTS
```{r echo = FALSE, message = FALSE, fig.height = 10, fig.width = 10}
grid.arrange(p1, p2, nrow = 1)
```

#### UMAP PLOTS
```{r echo = FALSE, message = FALSE, fig.height = 10, fig.width = 10}
grid.arrange(p3, p4, nrow = 1)

```


```{r echo = FALSE, message = FALSE, fig.height = 10, fig.width = 10}
# PCA PLOTS
# grid.arrange(p5, p6, nrow = 2)

```

Now we look at the cell id types in each sample. This bar plot shows the proportion of each type (cluster) in each sample.  
```{r echo = FALSE, message = FALSE}
barplot(height = freq_table)
```
