---
title: "Dimensionality reduction plots "
subtitle: '`r params$sample`'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  path_to_sce_norm: 'path_to_sce_norm'
  sce_red: 'sce_red'
  seed1: 'seed1'
  seed2: 'seed2'
  HVG: 'HVG'
  top_HVG: 'top_HVG'
  top_PCs: 'top_PCs'
  dm_path: 'dm_path'
  compute_dm: 'compute_dm'
  
---

This report includes a few figures below that can help you decide better how many PCs to include in further downstream analysis. 

```{r include = FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))

```
  
  
  
```{r message = FALSE}

# load sce
sce_norm <- readRDS(params$path_to_sce_norm)

# compute the number of PCs to use 
if (params$top_PCs == "computed") { # pipeline decides on the number of PCs to use 
  
  set.seed(params$seed1)
  dec.sce <- modelGeneVarByPoisson(sce_norm) # model gene variation, decompose it into technical and biological components
  
  # choose the HVG
  fraction_HGV <- as.numeric(params$top_HVG) / 100
  top.sce <- getTopHVGs(dec.sce, prop = fraction_HGV)
  
  denoised.sce <- denoisePCA(sce_norm, technical = dec.sce)
  PCs <- ncol(reducedDim(denoised.sce))
  
} else { # user decided how many PCs to use 
  
  top_PCs = as.numeric(params$top_PCs)
  PCs = top_PCs # number of PCs to use is given by the user in the top_PCs parameter
}

# deal with HVG
if (params$HVG == "yes") { # user wants to subset the sce to HVG 
  
  # pick the HVG
  gene_var <- modelGeneVar(sce_norm)
  
  fraction_HGV <- as.numeric(params$top_HVG) / 100
  chosen <- getTopHVGs(gene_var, prop = fraction_HGV)
  
  # subset to HVG
  sce_qc_hvg <- sce_norm[chosen,]
  
  # add the original count matrix with non HVG to our new sce
  altExp(sce_qc_hvg, "original") <- sce_norm
  sce_norm <- sce_qc_hvg
  
  set.seed(as.numeric(params$seed2))
  sce_norm <- runPCA(sce_norm, exprs_values = "logcounts", ncomponents = PCs)
  
} else { # compute PCA (and other dim reduction methods) without subsetting to top HVG
  
  set.seed(as.numeric(params$seed2))
  sce_norm <- runPCA(sce_norm, exprs_values = "logcounts", ncomponents = PCs)
  
}

# TSNE and UMAP 
set.seed(params$seed2)
sce_norm <- runTSNE(sce_norm, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, perplexity = 10)
sce_norm <- runUMAP(sce_norm, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, min_dist = 0.5, n_neighbors = 15, metric = "euclidean")


```

```{r echo = FALSE, message = FALSE, warning = FALSE}

if (params$compute_dm == "yes"){
  
  message(print("Computing diffusion map coordinates. This may take a while depending on how large your SCE object is."))

  sce_logcounts <- logcounts(sce_norm)  # access log-transformed counts matrix
  sce_logcounts <- as.matrix(sce_logcounts)
  
  # Make a diffusion map, this step can take a while depending on how large the sce object is, up to 10 minutes. 
  dm <- DiffusionMap(t(sce_logcounts), n_pcs = 30)
  
  # save this object separately from the sce 
  saveRDS(dm, params$dm_path)

}

```

```{r include = FALSE}
# save the output 
saveRDS(sce_norm, file = params$sce_red)

```

```{r include = FALSE}

# some seurat plots here to decide how many PCs to include 
# convert sce to seurat object
sobject <- as.Seurat(sce_norm, counts = "counts", data = "logcounts")
Idents(sobject) <- paste(unique(sce_norm$id))

# sobject@reductions$pca <- sobject@reductions$PCA
# sobject@reductions$tsne <- sobject@reductions$TSNE
# sobject@reductions$umap <- sobject@reductions$UMAP

sobject <- ScaleData(sobject)
sobject <- FindVariableFeatures(sobject)

set.seed(as.numeric(params$seed2))
sobject <- RunPCA(sobject, features = VariableFeatures(object = sobject))

```

#### Visualize top genes associated with reduction components, in top 6 PCs  
This figure below breaks down the variation in each of the top 6 PCs and shows which genes contribute most to overall variation in each PC. This could be helpful to decide on interesting genes for further downstream analysis. If you chose to retain ribosomal and mitochondrial genes in the previous steps of the pipeline, it's highly likely that they will contain a significant portion of the variance. 

```{r, fig.width = 14, fig.height = 10}

VizDimLoadings(sobject, dims = 1:6, reduction = "pca", ncol = 3)

```

### Deciding how many PCs to include in further analysis  
In the following heatmaps top genes contributing to each PC are visualized. Both cells and genes are ordered by their PC scores, and the 15 genes with the highest and 15 genes with the lowest PC scores are displayed for each PC. Note that the cells are given in the columns and genes are show in the rows. Here the goal is to pick the PCs that can divide the cell population into distinct groups. If all of the cells are the same color (pink), then that PC isn't as significant. 

```{r fig.width = 16, fig.height = 25}

DimHeatmap(sobject, dims = 1:15, cells = 500, balanced = TRUE)

```

### Finding significant PCs
‘Significant’ PCs will show a strong enrichment of genes with low p-values. To generate the plot below, a resampling test was implemented by the JackStraw procedure (Macosko et al). A subset of the data (1% by default) ws randomly permuted and PCA was computed again. This constructs a 'null distribution' of feature scores where the procedure was repeated numerous times. 'Significant' PCs were identified as those who have a strong enrichment of low p-value features.

```{r fig.width = 10, fig.height = 7, warning = FALSE, message=FALSE}

sobject <- JackStraw(object = sobject, reduction = "pca", dims = 20, num.replicate = 100,  prop.freq = 0.1, verbose = FALSE)

sobject <- ScoreJackStraw(object = sobject, dims = 1:20, reduction = "pca")
JackStrawPlot(object = sobject, dims = 1:20, reduction = "pca")


```

To further validate the results of the plot above, we can also check out contribution of each PC to total standard deviation and find the elbow point where the contribution to overall variation from each PC drops significantly. 

```{r fig.width = 10, fig.height = 4}

ElbowPlot(object = sobject)

```














