---
title: " `r params$ids` Summary stats and quality control results"
author: "asli munzur"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  id: 'DH16'
  ids: 'DH15-DH16'
---

```{r include = FALSE}
# how do i run this script? 
# you can run your samples through my entire pipeline using snakemake all the way from making an sce.
# or you can also save your data to my folders, and just run it - make sure to start my DH_organoid Rproj. 
# if you choose this methods, here is what you need to do: 
# make a directory with your sample ids in the following folders: data/processed, data/qc, data/normalized
# naming should be like this for the respective folders: 'sce.rds' 'sce_qc.rds' 'sce_norm.rds'
```

```{r include = FALSE}
# source the files for the functions we need 
library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
source(here('pipeline', 'summary_stats', 'qcPlots.R')) # has summary stats specific functions

```

```{r include = FALSE}

#extract the name_list from the parameters. we use the name list later on while making graphs. 
ids <- strsplit(params$ids, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
dh_organoid_ids <- list(ids[[1]], ids[[2]])
```

```{r include = FALSE}
# here we do a bit or renaming and preparation for what is to come. 
# find the directories where the sces are and load them
sce_raws <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'processed', id, 'sce.rds')))
sce_qcs <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'qc', id, 'sce_qc.rds')))

# rename the sces
names(sce_raws) <- dh_organoid_ids
names(sce_qcs) <- dh_organoid_ids

seurat_raws <- lapply(sce_raws, function(sce) CreateSeuratObject(counts = counts(sce), project = sce$id))
seurat_qc <- lapply(sce_qcs, function(sce) CreateSeuratObject(counts = counts(sce), project = sce$id))

# the QC metrics are saved in sces. now extract them from sces and load them into objects 
df_raws <- lapply(sce_raws, function(sce) colData(sce)) # this one has metrics before removing low quality cells 
df_qcs <- lapply(sce_qcs, function(sce) colData(sce)) # this one has metrics after removing low quality cells 

#rename the dfs 
names(df_raws) <- dh_organoid_ids
names(df_qcs) <- dh_organoid_ids

```

```{r include = FALSE}

all_raw <- merge(seurat_raws[[1]], seurat_raws[[2]], add.cell.ids = c(dh_organoid_ids[[1]], dh_organoid_ids[[2]]))
all_qc <- merge(seurat_qc[[1]], seurat_qc[[2]], add.cell.ids = c(dh_organoid_ids[[1]], dh_organoid_ids[[2]]))

all_raw[["percent.mito"]] <- PercentageFeatureSet(all_raw, pattern = "^MT-")
all_qc[["percent.mito"]] <- PercentageFeatureSet(all_qc, pattern = "^MT-")

all_raw[["percent_ribo"]] <- PercentageFeatureSet(all_raw, pattern = "^RP[SL]")
all_qc[["percent_ribo"]] <- PercentageFeatureSet(all_qc, pattern = "^RP[SL]")

feats <- c("nFeature_RNA","nCount_RNA","percent.mito","percent_ribo")

```

#### BEFORE QUALITY CONTROL
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 13}

VlnPlot(all_raw, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4) 

```

#### AFTER QUALITY CONTROL
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 13}

VlnPlot(all_qc, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4)

```

We can also visualize the information given above in scatter plots to better understand how they align, as shown below. 

#### BEFORE QUALITY CONTROL 

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

cowplot::plot_grid(ncol = 4,
  FeatureScatter(all_raw, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_mito", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_ribo", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_ribo", "percent_mito", group.by = "orig.ident", pt.size = .5)
)

```
#### AFTER QUALITY CONTROL 

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

cowplot::plot_grid(ncol = 4,
  FeatureScatter(all_qc, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_mito", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_ribo", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_ribo", "percent_mito", group.by = "orig.ident", pt.size = .5)
)

```

### Dimensionality reduction plots 

Next we include some dimensionality reduction plots. Plots on the left are before quality control, plots on the right are after. 
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

all_raw <- FindVariableFeatures(all_raw)
all_raw <- ScaleData(all_raw)

all_raw <- RunPCA(all_raw, verbose = FALSE)
all_raw <- RunUMAP(all_raw, dims = 1:30)
all_raw <- RunTSNE(all_raw, dims = 1:30)

p1 <- DimPlot(all_raw, reduction = 'pca', group.by = 'orig.ident', combine = FALSE)
p2 <- DimPlot(all_raw, reduction = 'tsne', group.by = 'orig.ident', combine = FALSE)
p3 <- DimPlot(all_raw, reduction = 'umap', group.by = 'orig.ident', combine = FALSE)

all_qc <- FindVariableFeatures(all_qc)
all_qc <- ScaleData(all_qc)

all_qc <- RunPCA(all_qc, verbose = FALSE)
all_qc <- RunUMAP(all_qc, dims = 1:30)
all_qc <- RunTSNE(all_qc, dims = 1:30)

p4 <- DimPlot(all_qc, reduction = 'pca', group.by = 'orig.ident', combine = FALSE)
p5 <- DimPlot(all_qc, reduction = 'tsne', group.by = 'orig.ident', combine = FALSE)
p6 <- DimPlot(all_qc, reduction = 'umap', group.by = 'orig.ident', combine = FALSE)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p1[[1]]), ggplotify::as.grob(p4[[1]]), nrow = 1)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p2[[1]]), ggplotify::as.grob(p5[[1]]), nrow = 1)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p3[[1]]), ggplotify::as.grob(p5[[1]]), nrow = 1)

```







