---
title: " `r params$ids` Summary stats and quality control results"
author: "asli munzur"
date: "11/5/2020"
output: 
  html_document:
    code_folding: hide

params:
  id: 'id'
  ids: 'pair_ids'
---

```{r include = FALSE}
# how do i run this script? 
# you can run your samples through my entire pipeline using snakemake all the way from making an sce.
# or you can also save your data to my folders, and just run it - make sure to start my DH_organoid Rproj. 
# if you choose this methods, here is what you need to do: 
# make a directory with your sample ids in the following folders: data/processed, data/qc, data/normalized
# naming should be like this for the respective folders: 'sce.rds' 'sce_qc.rds' 'sce_norm.rds'
```

```{r include = FALSE}
# source the files for the functions we need 
library(here)
source(here('..', 'sourceFiles', 'utilities.R')) # has all the packages etc and basic functions we need 
source(here('pipeline', 'summary_stats', 'qcPlots.R')) # has summary stats specific functions

```

```{r include = FALSE}

#extract the name_list from the parameters. we use the name list later on while making graphs. 
ids <- strsplit(params$ids, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
dh_organoid_ids <- list(ids[[1]], ids[[2]])
```

```{r include = FALSE}
# here we do a bit or renaming and preparation for what is to come. 
# find the directories where the sces are and load them
sce_raws <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'processed', id, 'sce.rds')))
sce_qcs <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'qc', id, 'sce_qc.rds')))

# rename the sces
names(sce_raws) <- dh_organoid_ids
names(sce_qcs) <- dh_organoid_ids

# the QC metrics are saved in sces. now extract them from sces and load them into objects 
df_raws <- lapply(sce_raws, function(sce) colData(sce)) # this one has metrics before removing low quality cells 
df_qcs <- lapply(sce_qcs, function(sce) colData(sce)) # this one has metrics after removing loq quality cells 

#rename the dfs 
names(df_raws) <- dh_organoid_ids
names(df_qcs) <- dh_organoid_ids

```


```{r include = FALSE}
# here we define a function to make plots. super simple function 
# data is the data we want to make graph of 
# what is what you are plotting on the y axis, the metric we are mainly interested in 
# whichcolor allows you to pick a color since each metric has its own color 
# name_list is the names of the datasets you use, we use it to name our plots. it must be list obviously 
# theme is the theme. i like using theme_amunzur
# yName is how you wanna name your y axis. we also use it to name our plots 
# ymax is the the limit on the y axis. 

makeQCplot <- function(data, what, whichcolor, name_list, theme, yName, yMax) {
  
  plot <- ggplot(data, aes(x = dataset, y = what)) +
    geom_bar(stat = 'identity', fill = whichcolor, color = "black") +
    ylab(yName) +
    xlab("dataset") +
    ylim(0, yMax) +
    ggtitle(yName) + 
    theme
  
  return(plot)
}
```

```{r include = FALSE}
# here we write a function to help us make dataframes that good for making plots later on 

makeTable <- function(df, index) {
  
  # first we must go from S4 to data frame to be able to use dplyr
  df <- as.data.frame(df) %>% 
    
    # get the mean for each column we are interested in 
    summarize(sum = as.integer(mean(sum)), detected = as.integer(mean(detected)), subsets_mito_sum = as.integer(mean(subsets_mito_sum)), 
              subsets_mito_percent =  as.integer(mean(subsets_mito_percent))) %>% 

    mutate(id = dh_organoid_ids[[index]]) %>% 

    # add a new column with the number of cells 
    mutate(cell_number = nrow(df)) %>% 
    
    # here we take advantage of the select function to reorder the columns 
    dplyr::select(id, cell_number, sum, detected, subsets_mito_sum, subsets_mito_percent) %>% 
    
    # and here we just rename them for clarity 
    rename(RNA_reads = sum, detected_genes = detected, mito_reads = subsets_mito_sum, mito_percent = subsets_mito_percent)
  
  return(df)
  
} # end of function 
```

```{r echo = FALSE, warning = FALSE}
# we gotta add some tables here before we jump into graphs. 
# we first make a very very basic table, then make it look better. 

BEFOREcontrolTable <- makeTable(df_raws[[1]], 1)
BEFOREexpTable <- makeTable(df_raws[[2]], 2)

# combine data from control and treatment in one table 
BEFOREtable <- rbind(BEFOREcontrolTable, BEFOREexpTable)

AFTERcontrolTable <- makeTable(df_qcs[[1]], 1)
AFTERexpTable <- makeTable(df_qcs[[2]], 2)

# combine data from control and treatment in one table 
AFTERtable <- rbind(AFTERcontrolTable, AFTERexpTable)

```

```{r echo = FALSE, warning = FALSE, results = 'asis'}
# here we just make the tables look nice and add a caption as well. 
BEFOREtable %>% 
  knitr::kable(caption = 'Table 1: Summary statistics showing cell number and mean values per cell- BEFORE quality control')
```


```{r echo = FALSE, warning = FALSE, results = 'asis'}
AFTERtable %>% 
  knitr::kable(caption = 'Table 2: Summary statistics showing cell number and mean values per cell- AFTER quality control')
```

## BASIC QC METRICS  
**These plots show these metrics:**  
- Mean number of RNA reads per cell  
- Mean number of detected genes per cell  
- Mean percentage of mitochondrial genes per cell  

```{r echo = FALSE, warning = FALSE, fig.height = 5, fig.width = 10}

# the following plots plot the MEAN values per cell 
# all 6 plots follow the same convention, easy to comprehend! 

# but before we do these plots, we will find the best number ymax value for each type of graph. 
# this value must stay the same before and after qc so that the basis of comparison stays the same. 
# for each graph we make, we will pick the higher value of the 4 values we have. 2 for each: before and 
# after qc. 

ymax_library <- max(mean(df_raws[[1]]$sum), 
                    mean(df_qcs[[1]]$sum), 
                    mean(df_raws[[2]]$sum), 
                    mean(df_qcs[[2]]$sum))

ymax_detected <- max(mean(df_raws[[1]]$detected), 
                    mean(df_qcs[[1]]$detected), 
                    mean(df_raws[[2]]$detected), 
                    mean(df_qcs[[2]]$detected))

ymax_mito <- max(mean(df_raws[[1]]$subsets_mito_percent), 
                    mean(df_qcs[[1]]$subsets_mito_percent), 
                    mean(df_raws[[2]]$subsets_mito_percent), 
                    mean(df_qcs[[2]]$subsets_mito_percent))


# the problem is these values are ugly af, lets make them look nicer in a rounding function: 
# actually this function doesnt work, so we will just use the ceiling function to round numbers. 
roundUpNice <- function(x, nice=c(1,2,4,5,6,8,10)) {
    if(length(x) != 1) stop("'x' must be of length 1")
    10^floor(log10(x)) * nice[[which(x <= 10^floor(log10(x)) * nice)[[1]]]]
}

# now update our y limits:
ymax_library <- ceiling(ymax_library)
ymax_detected <- ceiling(ymax_detected)
ymax_mito <- ceiling(ymax_mito)

# MAKE PER CELL PLOTS - BEFORE QC ####
# we will use the plots here for our before and after plotting & comparison

# compare LIBRARY SIZES - before 
dataset1 <- list(mean(df_raws[[1]]$sum), mean(df_raws[[2]]$sum))

data1 <- data.frame(dataset = unlist(dh_organoid_ids), 
           meanLibsize = unlist(dataset1))

plot1 <- makeQCplot(data1, data1$meanLibsize, 'chartreuse3', dh_organoid_ids, theme_amunzur, 'mean RNA reads', ymax_library)


# compare number of DETECTED - before 
dataset2 <- list(mean(df_raws[[1]]$detected ), mean(df_raws[[2]]$detected))

data2 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meandetected = unlist(dataset2))

plot2 <- makeQCplot(data2, data2$meandetected, 'brown3', dh_organoid_ids, theme_amunzur, 'mean detected genes', ymax_detected)



# compare MITO PERCENTAGE - before 
dataset3 <- list(mean(df_raws[[1]]$subsets_mito_percent ), mean(df_raws[[2]]$subsets_mito_percent))

data3 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meanmitoPercent = unlist(dataset3))

plot3 <- makeQCplot(data3, data3$meanmitoPercent, 'deeppink1', dh_organoid_ids, theme_amunzur, 'mean mito percent', ymax_mito)

grid.arrange(plot1, plot2, plot3, nrow = 1)
```

### AFTER QUALITY CONTROL  
```{r echo = FALSE, warning = FALSE, fig.height = 5, fig.width = 10}

# MAKE PER CELL PLOTS - AFTER QC ####
# we will use the plots here for our before and after plotting & comparison

# compare LIBRARY SIZES - after 
dataset4 <- list(mean(df_qcs[[1]]$sum), mean(df_qcs[[2]]$sum))

data4 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meanLibsize = unlist(dataset4))

plot4 <- makeQCplot(data4, data4$meanLibsize, 'chartreuse3', dh_organoid_ids, theme_amunzur, 'mean RNA reads', ymax_library)


# compare number of DETECTED - after 
dataset5 <- list(mean(df_qcs[[1]]$detected ), mean(df_qcs[[2]]$detected))

data5 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meandetected = unlist(dataset5))

plot5 <- makeQCplot(data5, data5$meandetected, 'brown3', dh_organoid_ids, theme_amunzur, 'mean detected genes', ymax_detected)



# compare MITO PERCENTAGE - after  
dataset6 <- list(mean(df_qcs[[1]]$subsets_mito_percent ), mean(df_qcs[[2]]$subsets_mito_percent))

data6 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meanmitoPercent = unlist(dataset6))

plot6 <- makeQCplot(data6, data6$meanmitoPercent, 'deeppink1', dh_organoid_ids, theme_amunzur, 'mean mito percent', ymax_mito)


# show all 6 plots in one panel 
grid.arrange(plot4, plot5, plot6, nrow = 1)
```

### Let's look at after QC results a little more closely: 
```{r echo = FALSE, warning = FALSE, fig.height = 5, fig.width = 10}

ymax_library <- ceiling(max(mean(df_qcs[[1]]$sum), 
                    mean(df_qcs[[2]]$sum)))

ymax_detected <- ceiling(max(mean(df_qcs[[1]]$detected), 
                    mean(df_qcs[[2]]$detected)))

ymax_mito <- ceiling(max(mean(df_qcs[[1]]$subsets_mito_percent), 
                    mean(df_qcs[[2]]$subsets_mito_percent)))

# MAKE PER CELL PLOTS - AFTER QC ####

# compare LIBRARY SIZES - after 
dataset4 <- list(mean(df_qcs[[1]]$sum), mean(df_qcs[[2]]$sum))

data4 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meanLibsize = unlist(dataset4))

plot4 <- makeQCplot(data4, data4$meanLibsize, 'chartreuse3', dh_organoid_ids, theme_amunzur, 'mean RNA reads', ymax_library)


# compare number of DETECTED - after 
dataset5 <- list(mean(df_qcs[[1]]$detected ), mean(df_qcs[[2]]$sum))

data5 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meandetected = unlist(dataset5))

plot5 <- makeQCplot(data5, data5$meandetected, 'brown3', dh_organoid_ids, theme_amunzur, 'mean detected genes', ymax_detected)



# compare MITO PERCENTAGE - after  
dataset6 <- list(mean(df_qcs[[1]]$subsets_mito_percent ), mean(df_qcs[[2]]$subsets_mito_percent))

data6 <- data.frame(dataset = unlist(dh_organoid_ids), 
                   meanmitoPercent = unlist(dataset6))

plot6 <- makeQCplot(data6, data6$meanmitoPercent, 'deeppink1', dh_organoid_ids, theme_amunzur, 'mean mito percent', ymax_mito)


# show all plots in one panel 
grid.arrange(plot4, plot5, plot6, nrow = 1)
```



```{r include = FALSE}
# MAKE TOTAL GRAPHS - not per cell - BEFORE ####

# lib size 
before1 <-  totalCountsPlot(as.data.frame(df_raws[[1]]), ymax_library) + ggtitle(dh_organoid_ids[[1]])
before2 <- totalCountsPlot(as.data.frame(df_raws[[2]]), ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
before3 <- geneNumberPlot(as.data.frame(df_raws[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
before4 <- geneNumberPlot(as.data.frame(df_raws[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
before5 <- MitoPercentPlot(as.data.frame(df_raws[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
before6 <- MitoPercentPlot(as.data.frame(df_raws[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])

# UMI and genes 
before7 <- UMIandGenesPlot(as.data.frame(df_raws[[1]])) + ggtitle(dh_organoid_ids[[1]])
before8 <- UMIandGenesPlot(as.data.frame(df_raws[[2]])) + ggtitle(dh_organoid_ids[[2]])

# MAKE TOTAL GRAPHS - not per cell - AFTER ####

before9 <-  totalCountsPlot(as.data.frame(df_qcs[[1]]), ymax_library) + ggtitle(dh_organoid_ids[[1]])
before10 <- totalCountsPlot(as.data.frame(df_qcs[[2]]), ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
before11 <- geneNumberPlot(as.data.frame(df_qcs[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
before12 <- geneNumberPlot(as.data.frame(df_qcs[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
before13 <- MitoPercentPlot(as.data.frame(df_qcs[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
before14 <- MitoPercentPlot(as.data.frame(df_qcs[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])

# UMI and genes 
before15 <- UMIandGenesPlot(as.data.frame(df_qcs[[1]])) + ggtitle(dh_organoid_ids[[1]])
before16 <- UMIandGenesPlot(as.data.frame(df_qcs[[2]])) + ggtitle(dh_organoid_ids[[2]])
```



```{r include = FALSE}

# okay now lets do the same finding the y limit method for the histograms: 


# find the max value from 4 sets and make that the y limit: 
max1 <- max(ggplot_build(before1)$data[[1]]$count)
max2 <- max(ggplot_build(before2)$data[[1]]$count)
max3 <- max(ggplot_build(before9)$data[[1]]$count)
max4 <- max(ggplot_build(before10)$data[[1]]$count)

ymax_library <- max(max1, max2, max3, max4)

max1 <- max(ggplot_build(before3)$data[[1]]$count)
max2 <- max(ggplot_build(before4)$data[[1]]$count)
max3 <- max(ggplot_build(before11)$data[[1]]$count)
max4 <- max(ggplot_build(before12)$data[[1]]$count)

ymax_detected <- max(max1, max2, max3, max4)

max1 <- max(ggplot_build(before5)$data[[1]]$count)
max2 <- max(ggplot_build(before6)$data[[1]]$count)
max3 <- max(ggplot_build(before13)$data[[1]]$count)
max4 <- max(ggplot_build(before14)$data[[1]]$count)

ymax_mito <-max(max1, max2, max3, max4)
```


```{r include = FALSE}
# MAKE TOTAL GRAPHS - not per cell - BEFORE ####

# lib size 
before1 <-  totalCountsPlot(as.data.frame(df_raws[[1]]), ymax_library) + ggtitle(dh_organoid_ids[[1]])
before2 <- totalCountsPlot(as.data.frame(df_raws[[2]]), ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
before3 <- geneNumberPlot(as.data.frame(df_raws[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
before4 <- geneNumberPlot(as.data.frame(df_raws[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
before5 <- MitoPercentPlot(as.data.frame(df_raws[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
before6 <- MitoPercentPlot(as.data.frame(df_raws[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])

# UMI and genes 
before7 <- UMIandGenesPlot(as.data.frame(df_raws[[1]])) + ggtitle(dh_organoid_ids[[1]])
before8 <- UMIandGenesPlot(as.data.frame(df_raws[[2]])) + ggtitle(dh_organoid_ids[[2]])

# MAKE TOTAL GRAPHS - not per cell - AFTER ####

before9 <-  totalCountsPlot(as.data.frame(df_qcs[[1]]), ymax_library) + ggtitle(dh_organoid_ids[[1]])
before10 <- totalCountsPlot(as.data.frame(df_qcs[[2]]), ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
before11 <- geneNumberPlot(as.data.frame(df_qcs[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
before12 <- geneNumberPlot(as.data.frame(df_qcs[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
before13 <- MitoPercentPlot(as.data.frame(df_qcs[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
before14 <- MitoPercentPlot(as.data.frame(df_qcs[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])

# UMI and genes 
before15 <- UMIandGenesPlot(as.data.frame(df_qcs[[1]])) + ggtitle(dh_organoid_ids[[1]])
before16 <- UMIandGenesPlot(as.data.frame(df_qcs[[2]])) + ggtitle(dh_organoid_ids[[2]])
```


```{r include = FALSE}
# lets make some after graphs by zooming in a little bit: 
# so now we need to find the ymax value among the df_qcs only.

after1 <- totalCountsPlot(as.data.frame(df_qcs[[1]]),  ymax_library) + ggtitle(dh_organoid_ids[[1]])
after2 <- totalCountsPlot(as.data.frame(df_qcs[[2]]),  ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
after3 <- geneNumberPlot(as.data.frame(df_qcs[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
after4 <- geneNumberPlot(as.data.frame(df_qcs[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
after5 <- MitoPercentPlot(as.data.frame(df_qcs[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
after6 <- MitoPercentPlot(as.data.frame(df_qcs[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])

# updatde them with the correct ymax values: 
ymax_library <- max(max(ggplot_build(after1)$data[[1]]$count), max(ggplot_build(after2)$data[[1]]$count))
ymax_detected <- max(max(ggplot_build(after3)$data[[1]]$count), max(ggplot_build(after4)$data[[1]]$count))
ymax_mito <- max(max(ggplot_build(after5)$data[[1]]$count), max(ggplot_build(after6)$data[[1]]$count))

# make those graphs again: 
after1 <-  totalCountsPlot(as.data.frame(df_qcs[[1]]), ymax_library) + ggtitle(dh_organoid_ids[[1]])
after2 <- totalCountsPlot(as.data.frame(df_qcs[[2]]), ymax_library) + ggtitle(dh_organoid_ids[[2]])

# detected genes 
after3 <- geneNumberPlot(as.data.frame(df_qcs[[1]]), ymax_detected) + ggtitle(dh_organoid_ids[[1]])
after4 <- geneNumberPlot(as.data.frame(df_qcs[[2]]), ymax_detected) + ggtitle(dh_organoid_ids[[2]])

# mito percent 
after5 <- MitoPercentPlot(as.data.frame(df_qcs[[1]]), ymax_mito) + ggtitle(dh_organoid_ids[[1]])
after6 <- MitoPercentPlot(as.data.frame(df_qcs[[2]]), ymax_mito) + ggtitle(dh_organoid_ids[[2]])
```

These following graphs show the distribution of quality control metrics across all cells. 
Red line shows the mean. 

### Number of RNA reads  
##### before QC 
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize these graphs 
gridExtra::grid.arrange(before1, before2, nrow = 1)
```

##### after QC 
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 10}
# now we visualize these graphs 
gridExtra::grid.arrange(before9, before10, after1, after2, nrow = 1)
```


### Number of detected genes  
##### before QC
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize these graphs 
gridExtra::grid.arrange(before3, before4, nrow = 1)
```

##### after QC 
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 10}
# now we visualize these graphs 
gridExtra::grid.arrange(before11, before12, after3, after4, nrow = 1)
```


### Mitochondrial gene percentage  
##### before QC
Blue line shows 10% threshold.  
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize these graphs 
gridExtra::grid.arrange(before5, before6, nrow = 1)
```


##### after QC
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 10}
# now we visualize these graphs 
gridExtra::grid.arrange(before13, before14, after5, after6, nrow = 1)
```


### Comparing number of RNA reads and number of detected genes 
##### before QC

```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize these graphs 
gridExtra::grid.arrange(before7, before8, nrow = 1)
```

##### after QC

```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize these graphs 
gridExtra::grid.arrange(before15, before16, nrow = 1)
```


```{r include = FALSE}
# some cool plots with scater:
# CONTROL - DH25_control
control_raw_withMetrics <- calculateQCMetrics(sce_raws[[1]])
control_coolplot1 <- plotExprsFreqVsMean(control_raw_withMetrics) + ggtitle(paste(dh_organoid_ids[[1]], 'before QC', sep = '-'))

control_qc_withMetrics <- calculateQCMetrics(sce_qcs[[1]])
control_coolplot2 <- plotExprsFreqVsMean(control_qc_withMetrics) + ggtitle(paste(dh_organoid_ids[[1]], 'after QC', sep = '-'))

# EXPERIMENTAL - DH25
exp_raw_withMetrics <- calculateQCMetrics(sce_raws[[2]])
exp_coolplot1 <- plotExprsFreqVsMean(exp_raw_withMetrics) + ggtitle(paste(dh_organoid_ids[[2]], 'before QC', sep = '-'))

exp_qc_withMetrics <- calculateQCMetrics(sce_qcs[[2]])
exp_coolplot2 <- plotExprsFreqVsMean(exp_qc_withMetrics) + ggtitle(paste(dh_organoid_ids[[2]], 'after QC', sep = '-'))

```

Frequency of expression (i.e., number of cells with non-zero expression) plotted against the mean.  
These two metrics should be positively correlated with each other for most genes.

BEFORE QC
```{r echo = FALSE, warning = FALSE, fig.height = 10, fig.width = 10}
gridExtra::grid.arrange(control_coolplot1, exp_coolplot1, nrow = 1)
```

AFTER QC
```{r echo = FALSE, warning = FALSE, fig.height = 10, fig.width = 10}
gridExtra::grid.arrange(control_coolplot2, exp_coolplot2)
```


### REDUCED DIMENSIONALITY PLOTS 
Here we include the reduced dimensionality plots before and after batch normalization.  
Note that these plots are only for the cells that underwent quality control.  

```{r include = FALSE}
# BEFORE BATCH NORM 

# here, we need to use the normalized sces to run dim reduction. so lets load them and make a list. 
sce_norms <- lapply(dh_organoid_ids, function(id) readRDS(here('data', 'normalized', id, 'sce_norm.rds')))

# remove the reduced dimensions 
sce_norms <- lapply(sce_norms, function(sce){
      for(n in reducedDimNames(sce)){
        reducedDim(sce, n) <- NULL
      }
      return(sce)
    })

# combine sces
sce <- combine_sces(sce_list = sce_norms, prefix_col_name = 'id', prefix_cell_name = TRUE)
  
# Run UMAP, TSNE, PCA for visualization downstream steps: clustering, cellassign, ect.
sce <- runPCA(sce, exprs_values = "logcounts", ncomponents = 200)
sce <- runTSNE(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3)
sce <- runUMAP(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3,
                    min_dist = 0.01, n_neighbors = 15, metric = "euclidean")

# make the reduced dims plots 
PCA_noBatchNorm <- plotReducedDim(sce, dimred = 'PCA') + ggtitle('before batch norm - PCA')
TSNE_noBatchNorm <- plotReducedDim(sce, dimred = 'TSNE') + ggtitle('before batch norm - TSNE')
UMAP_noBatchNorm <- plotReducedDim(sce, dimred = 'UMAP') + ggtitle('before batch norm - UMAP')

```


```{r include = FALSE}
# AFTER BATCH NORM

# here, we need to use the normalized sces to run dim reduction. so lets load them and make a list. 
sce_norms <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'normalized', id, 'sce_norm.rds')))

# batch normalization 
sce_norms <- scran_batch_norm(sce_norms)

# combine sces
sce <- combine_sces(sce_list = sce_norms, prefix_col_name = 'id', prefix_cell_name = TRUE)
  
# Run UMAP, TSNE, PCA for visualization downstream steps: clustering, cellassign, ect.
sce <- runPCA(sce, exprs_values = "logcounts", ncomponents = 200)
sce <- runTSNE(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3)
sce <- runUMAP(sce, exprs_values = "logcounts", ntop = 500, ncomponents = 3,
                    min_dist = 0.01, n_neighbors = 15, metric = "euclidean")

# make the reduced dims plots
TSNEplot <- plotReducedDim(sce, 'TSNE') + ggtitle('after batch Norm - TSNE')
UMAPplot <- plotReducedDim(sce, 'UMAP') + ggtitle('after batch Norm - UMAP')
PCAplot <- plotReducedDim(sce, 'PCA') + ggtitle('after batch Norm - PCA')

```

TSNE
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize
gridExtra::grid.arrange(TSNE_noBatchNorm, TSNEplot, nrow = 1)
```

UMAP
```{r echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 7}
# now we visualize
gridExtra::grid.arrange(UMAP_noBatchNorm, UMAPplot, nrow = 1)
```

### REPORTER GENE DETECTION 

```{r include = FALSE}

# find GFP 
find_reporter_genes('gfp', sce)







```






