---
title: " `r params$ids` Summary stats and quality control results"
subtitle: '`r params$pair_ids`'
author: "Asli Munzur"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  pair_ids: 'DH15-DH16' # to run this, write TWO ids here, separated by -
---

```{css, echo=FALSE}
    body .main-container {
      max-width: 90% !important;
      width: 90% !important;
    }

    body {
      max-width: 90% !important;
    }
```

```{r include = FALSE}
# source the files for the functions we need 
library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))

```

```{r include = FALSE}

#extract the name_list from the parameters. we use the name list later on while making graphs. 
ids <- strsplit(params$pair_ids, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
dh_organoid_ids <- list(ids[[1]], ids[[2]])
```

```{r include = FALSE}

# extract some information 
sce_raws <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'processed', id, 'sce.rds')))
sce_qcs <- lapply(dh_organoid_ids, function(id) readRDS(here('..', 'data', 'qc', id, 'sce_qc.rds')))

seurat_raws <- lapply(sce_raws, function(sce) CreateSeuratObject(counts = counts(sce), project = sce$id))
seurat_qc <- lapply(sce_qcs, function(sce) CreateSeuratObject(counts = counts(sce), project = sce$id))

df_raws <- lapply(sce_raws, function(sce) colData(sce)) # this one has metrics before removing low quality cells 
df_qcs <- lapply(sce_qcs, function(sce) colData(sce)) # this one has metrics after removing loq quality cells 

```

```{r include = FALSE}
# here we write a function to help us make dataframes that good for making plots later on 

makeTable <- function(df, index) {
  
  # first we must go from S4 to data frame to be able to use dplyr
  df <- as.data.frame(df) %>% 
    
    # get the mean for each column we are interested in 
    summarize(sum = as.integer(mean(sum)), detected = as.integer(mean(detected)), subsets_mito_sum = as.integer(mean(subsets_mito_sum)), 
              subsets_mito_percent =  as.integer(mean(subsets_mito_percent))) %>% 

    mutate(id = dh_organoid_ids[[index]]) %>% 

    # add a new column with the number of cells 
    mutate(cell_number = nrow(df)) %>% 
    
    # here we take advantage of the select function to reorder the columns 
    dplyr::select(id, cell_number, sum, detected, subsets_mito_sum, subsets_mito_percent) %>% 
    
    # and here we just rename them for clarity 
    rename(RNA_reads = sum, detected_genes = detected, mito_reads = subsets_mito_sum, mito_percent = subsets_mito_percent)
  
  return(df)
  
} # end of function 
```

```{r echo = FALSE, warning = FALSE}
# we gotta add some tables here before we jump into graphs. 
# we first make a very very basic table, then make it look better. 

BEFOREcontrolTable <- makeTable(df_raws[[1]], 1)
BEFOREexpTable <- makeTable(df_raws[[2]], 2)

# combine data from control and treatment in one table 
BEFOREtable <- rbind(BEFOREcontrolTable, BEFOREexpTable)

AFTERcontrolTable <- makeTable(df_qcs[[1]], 1)
AFTERexpTable <- makeTable(df_qcs[[2]], 2)

# combine data from control and treatment in one table 
AFTERtable <- rbind(AFTERcontrolTable, AFTERexpTable)

```

```{r echo = FALSE, warning = FALSE, results = 'asis'}
# here we just make the tables look nice and add a caption as well. 
BEFOREtable %>% 
  knitr::kable(caption = 'Table 1: Summary statistics showing cell number and mean values per cell- BEFORE quality control') %>% 
  kable_styling(full_width = F)
```


```{r echo = FALSE, warning = FALSE, results = 'asis'}
AFTERtable %>% 
  knitr::kable(caption = 'Table 2: Summary statistics showing cell number and mean values per cell- AFTER quality control') %>% 
  kable_styling(full_width = F)
```

```{r include = FALSE}

all_raw <- merge(seurat_raws[[1]], seurat_raws[[2]], add.cell.ids = c(dh_organoid_ids[[1]], dh_organoid_ids[[2]]))
all_qc <- merge(seurat_qc[[1]], seurat_qc[[2]], add.cell.ids = c(dh_organoid_ids[[1]], dh_organoid_ids[[2]]))

all_raw[["percent_mito"]] <- PercentageFeatureSet(all_raw, pattern = "^MT-")
all_qc[["percent_mito"]] <- PercentageFeatureSet(all_qc, pattern = "^MT-")

all_raw[["percent_ribo"]] <- PercentageFeatureSet(all_raw, pattern = "^RP[SL]")
all_qc[["percent_ribo"]] <- PercentageFeatureSet(all_qc, pattern = "^RP[SL]")

feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo")

```

#### BEFORE QUALITY CONTROL
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 13}

VlnPlot(all_raw, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4) 

```

#### AFTER QUALITY CONTROL
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 13}

VlnPlot(all_qc, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4)

```

We can also visualize the information given above in scatter plots to better understand how they align, as shown below. Pearson correlation between the two features is displayed above the plot. 

#### BEFORE QUALITY CONTROL 

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 17}

cowplot::plot_grid(ncol = 4,
  FeatureScatter(all_raw, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_mito", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_ribo", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_raw, "percent_ribo", "percent_mito", group.by = "orig.ident", pt.size = .5)
)

```
#### AFTER QUALITY CONTROL 

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 17}

cowplot::plot_grid(ncol = 4,
  FeatureScatter(all_qc, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_mito", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_ribo", "nFeature_RNA", group.by = "orig.ident", pt.size = .5),
  FeatureScatter(all_qc, "percent_ribo", "percent_mito", group.by = "orig.ident", pt.size = .5)
)

```

### Dimensionality reduction plots 

Next we include some dimensionality reduction plots. Plots on the left are before quality control, plots on the right are after. 
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

all_raw <- FindVariableFeatures(all_raw)
all_raw <- ScaleData(all_raw)

all_raw <- RunPCA(all_raw, verbose = FALSE)
all_raw <- RunUMAP(all_raw, dims = 1:30)
all_raw <- RunTSNE(all_raw, dims = 1:30, check_duplicates = FALSE)

p1 <- DimPlot(all_raw, reduction = 'pca', group.by = 'orig.ident', combine = FALSE)
p2 <- DimPlot(all_raw, reduction = 'tsne', group.by = 'orig.ident', combine = FALSE)
p3 <- DimPlot(all_raw, reduction = 'umap', group.by = 'orig.ident', combine = FALSE)

all_qc <- FindVariableFeatures(all_qc)
all_qc <- ScaleData(all_qc)

all_qc <- RunPCA(all_qc, verbose = FALSE)
all_qc <- RunUMAP(all_qc, dims = 1:30)
all_qc <- RunTSNE(all_qc, dims = 1:30, check_duplicates = FALSE)

p4 <- DimPlot(all_qc, reduction = 'pca', group.by = 'orig.ident', combine = FALSE)
p5 <- DimPlot(all_qc, reduction = 'tsne', group.by = 'orig.ident', combine = FALSE)
p6 <- DimPlot(all_qc, reduction = 'umap', group.by = 'orig.ident', combine = FALSE)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p1[[1]]), ggplotify::as.grob(p4[[1]]), nrow = 1)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p2[[1]]), ggplotify::as.grob(p5[[1]]), nrow = 1)

```

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 15}

grid.arrange(ggplotify::as.grob(p3[[1]]), ggplotify::as.grob(p5[[1]]), nrow = 1)

```

### Most variable features  
Here we find the 3000 most variable features (genes) in two datasets **after** quality control. Datatables below have include these genes, but we label the top 15 genes for both datasets. 

```{r include = FALSE}

seurat_qc[[1]] <- FindVariableFeatures(seurat_qc[[1]], selection.method = "vst", nfeatures = 3000)
seurat_qc[[2]] <- FindVariableFeatures(seurat_qc[[2]], selection.method = "vst", nfeatures = 3000)


# Identify the 10 most highly variable genes
top10_1 <- head(VariableFeatures(seurat_qc[[1]]), 15)
top10_2 <- head(VariableFeatures(seurat_qc[[2]]), 15)

# all the variable genes 
one_all <- VariableFeatures(seurat_qc[[1]])
two_all <- VariableFeatures(seurat_qc[[2]])


# plot variable features with labels
plot1 <- VariableFeaturePlot(seurat_qc[[1]])
plot2 <- LabelPoints(plot = plot1, points = top10_1, repel = TRUE)
plot2 <- plot2 + ggtitle(paste(dh_organoid_ids[[1]])) + theme(legend.position = "bottom")


plot3 <- VariableFeaturePlot(seurat_qc[[2]])
plot4 <- LabelPoints(plot = plot3, points = top10_2, repel = TRUE)
plot4 <- plot4 + ggtitle(paste(dh_organoid_ids[[2]])) + theme(legend.position = "bottom")



```


```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 10}

grid.arrange(plot2, plot4, nrow = 1)

```


#### Most variable genes in `r paste(dh_organoid_ids[[1]])`  
```{r}

DT::datatable(HVFInfo(seurat_qc[[1]])[VariableFeatures(seurat_qc[[1]]), ],
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip',
                 buttons = list("copy", "excel", "csv", "pdf")))


```


#### Most variable genes in `r paste(dh_organoid_ids[[2]])`  
```{r}

DT::datatable(HVFInfo(seurat_qc[[2]])[VariableFeatures(seurat_qc[[2]]), ],
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip',
                 buttons = list("copy", "excel", "csv", "pdf")))


```

