---
title: Correct for cell cycle effect
author: "Harper Cheng"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    
params:
  pair_ids: 'pair_ids'
  sce_cell_cycle: 'sce_cell_cycle'
---


## Motivation

scRNA-seq suffers from systematic bias caused by biological noises which could distort real biological signals. One of the major sources of biological noise is the cell cycle. Cells of the same cell type but are in different cell cycle may exhibit different expression profile. This will likely undermine the performance of clustering for cell type detection, e.g., cells might cluster by both cell type and cell cycle status. We will first assign cell cycle status to each cell and visualize the distribution. If PCA shows clusterings based on cell cycle status, we will then correct for the cell cycle effect. 


```{r setup, include=FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
set.seed(123)

# snakemake passes the input files as one string separated by a space. we need to divide them into separate lists: 
paths <- as.list(unlist(strsplit(params$sce_cell_cycle[[1]], ' '))) 

# load cell_cycle sce:
sces <- lapply(paths, function(some_path) readRDS(some_path)) 

ids <- strsplit(params$pair_ids, "-")[[1]] %>% as.list(c(unlist(ids)))

```


```{r}

### 1. Bar plot showing the proportion of cells belonging to each cell cycle phase

p <- list()

for (i in 1:length(sces)){
  if (is.null(sces[[i]]$cell_type)){
    
     # If no information on cell type:
     # Tabulate cell cycle results into a dataframe and plot: showing proportion of cell cycle phases within each cell type
    df <- data.frame(cell=sces[[i]]$Barcode, cell_cycle=sces[[i]]$cell_cycle)
    df <- df %>% dplyr::group_by(cell_cycle) %>% dplyr::summarize(count=n()) %>% dplyr::mutate(prop=count/sum(count))
    
     # Visualize the proportion of cell cycle phases for each sample.
    p[[i]] <- df %>% ggplot(aes(fill=cell_cycle, y=prop, x=cell_cycle)) +
      geom_bar(position="dodge", stat="identity") +
      geom_text(aes(label=round(prop,2)), size=3) +
      labs(x='Cell Cycle Phases', y='Proportion of Cells', title=ids[[i]])
    }
    else {
      
       # If plotting cell cycle for each cell type:
      df <- data.frame(cell=sces[[i]]$Barcode, cell_type=sces[[i]]$cell_type, cell_cycle=sces[[i]]$cell_cycle)
      df <- df %>% dplyr::group_by(cell_type, cell_cycle) %>% dplyr::summarize(count=n()) %>% dplyr::mutate(prop=count/sum(count))
  
      p[[i]] <- df %>% ggplot(aes(fill=cell_cycle, y=prop, x=cell_cycle)) +
        geom_bar(position="dodge", stat="identity") +
        facet_wrap(~cell_type) +
        geom_text(aes(label=round(prop,2)), size=3) +
        labs(x='Cell Cycle Phases', y='Proportion of Cells', title=ids[[i]])
    }
 }

```


```{r}

### 2. Correct for cell cycle effect and visualize in PCA space

# We can regress out the cell cycle effect and plot:
# one assumption is that the two subpopulations have similar cell cycle phase distribution;
# this can be examined by looking at the bar plots above.

pca <- tsne <- umap <- list()
sce_corrected <- list()

for(i in 1:length(sces)){
  
  # correct for cell cycle effect by regressing it out 
  reg.nocycle <- regressBatches(sces[[i]], batch=sces[[i]]$cell_cycle)
  reg.nocycle <- runPCA(reg.nocycle, exprs_values="corrected")
  reg.nocycle <- runTSNE(reg.nocycle, dimred="PCA", exprs_values="corrected")
  reg.nocycle <- runUMAP(reg.nocycle, dimred = "PCA", exprs_values="corrected")
  
  sce_corrected[[i]] <- reg.nocycle
  
  # PCA showing clustering before and after cell cycle correction
  pca[[i]] <- arrangeGrob(
    plotPCA(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": Before")),
    plotPCA(reg.nocycle, colour_by="batch") + ggtitle(paste0(ids[[i]], ": After")),
    ncol=2)
  
  # tSNE showing clustering
  tsne[[i]] <- arrangeGrob(
    plotTSNE(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": Before")),
    plotTSNE(reg.nocycle, colour_by="batch") + ggtitle(paste0(ids[[i]], ": After")),
    ncol=2)
    
  # UMAP showing clustering
  umap[[i]] <- arrangeGrob(
    plotUMAP(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": Before")),
    plotUMAP(reg.nocycle, colour_by="batch") + ggtitle(paste0(ids[[i]], ": After")),
    ncol=2)
}

```


## Bar plot of cell cycle phase

Bar plots showing distribution of cell cycle phases for each sample.

```{r}
plot_grid(p[[1]], p[[2]], ncol=2)
```



## PCA plot before and after correcting for cell cycle effect

We can regress out the cell cycle effect and plot cells on a PCA space. One assumption for regressing out cell cycle is that the two subpopulations have similar cell cycle phase distribution. This can be examined by looking at the bar plots above.

```{r}
plot_grid(pca[[1]], pca[[2]], nrow=2)
```

### PCA plot showing cell cycle clustering and cell clustering before cell cycle correction

Next, we plot PCA with cell cycle information alongside PCA showing cell clustering. This could inform us if cells cluster by cell cycle phases.

```{r}
clustering <- cycle <- list()

for(i in 1:length(sces)){
  cycle[[i]] <- plotPCA(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": cell cycle (PCA)"))
  
  clustering[[i]] <- plotReducedDim(sces[[i]], dimred="PCA", ncomponents = 2, colour_by="cluster") + 
    ggtitle(paste0(ids[[i]], ": cell clustering (PCA)"))
}

plot_grid(cycle[[1]], clustering[[1]], nrow=1)
plot_grid(cycle[[2]], clustering[[2]], nrow=1)

```



## tSNE before and after correcting for cell cycle effect

```{r}
plot_grid(tsne[[1]], tsne[[2]], nrow=2)
```


### tSNE showing cell cycle clustering and cell clustering before cell cycle correction

```{r}
clustering <- cycle <- list()

for(i in 1:length(sces)){
  cycle[[i]] <- plotTSNE(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": cell cycle (tSNE)"))
  
  clustering[[i]] <- plotReducedDim(sces[[i]], dimred="TSNE", ncomponents = 2, colour_by="cluster") + 
    ggtitle(paste0(ids[[i]], ": cell clustering (tSNE)"))
}

plot_grid(cycle[[1]], clustering[[1]], nrow=1)
plot_grid(cycle[[2]], clustering[[2]], nrow=1)

```



## UMAP before and after correting for cell cycle effect

```{r}
plot_grid(umap[[1]], umap[[2]], nrow=2)
```

### UMAP showing cell cycle clustering and cell clustering before cell cycle correction

```{r}
clustering <- cycle <- list()

for(i in 1:length(sces)){
  cycle[[i]] <- plotUMAP(sces[[i]], colour_by=I(sces[[i]]$cell_cycle)) + ggtitle(paste0(ids[[i]], ": cell cycle (UMAP)"))
  
  clustering[[i]] <- plotReducedDim(sces[[i]], dimred="UMAP", ncomponents = 2, colour_by="cluster") + 
    ggtitle(paste0(ids[[i]], ": cell clustering (UMAP)"))
}

plot_grid(cycle[[1]], clustering[[1]], nrow=1)
plot_grid(cycle[[2]], clustering[[2]], nrow=1)

```



## Bar plot of cell cycle phase for each cluster

```{r, echo=FALSE}

p <- list()

for (i in 1:length(sces)){
    
    df <- data.frame(cell=sces[[i]]$Barcode, cluster=sces[[i]]$cluster, cell_cycle=sces[[i]]$cell_cycle)
    df <- df %>% dplyr::group_by(cluster, cell_cycle) %>% dplyr::summarize(count=n()) %>% dplyr::mutate(prop=count/sum(count))
  
      p[[i]] <- df %>% ggplot(aes(fill=cell_cycle, y=prop, x=cell_cycle)) +
        geom_bar(position="dodge", stat="identity") +
        facet_wrap(~cluster) +
        geom_text(aes(label=round(prop,2)), size=3) +
        labs(x='Cell Cycle Phases', y='Proportion of Cells', title=paste0(ids[[i]], ": cell cycle phase distribution for each cluster"))

 }

p[[1]]
p[[2]]

```

