---
title: Cell assign results 
subtitle: '`r params$pair_ids` - `r params$id_type`'
author: "aslı"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  pair_ids: 'VOA11068_ENOC-DH13-DH18' # write here the ids you want to integrate, separate by '-'
  id_type: 'id_type' # here you can write a quick summary of what these pair_ids are, like tissue types etc. 
---

#### In this report we show the results of cell assign runs on the samples. We include some dim reduction plots with cell types and clustering information. We first show results of cell assign run on integrated sample, then we give the plots for individual tumours.   

**Cell assign is able to detect the following cell types:**  
* Blymphocyte  
* Tlymphocyte  
* CytotoxicT  
* Monocyte  
* Epithelial  
* Myofibroblast  
* Vascular Smooth Muscle  
* Endothelial  
* PlasmaB  
* Epithelial Ciliated  
* Endometrial Stem  
* Mesenchymal Stem  
* Tumor cells: PAX8, WT1, STMN1 (my addition)  
* Mesenchymal cells  

```{r include = FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
source(here('pipeline', 'cellassign', 'create_marker_matrix.R'))
source(here('pipeline', 'cellassign', 'assign_cell_type.R'))
```

```{r include = FALSE}

# get separate ids from the params$pair_ids
id.list <- strsplit(params$pair_ids, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
id.list <- as.list(c(unlist(list(id.list))))

```


```{r include = FALSE}
# we start with the plots for uncorrected samples  

uncorrected <- readRDS(here('..', 'data', 'integrated', params$pair_ids, 'uncorrected.rds'))

# some dim red plots 
p1 <- plotReducedDim(uncorrected, "TSNE", colour_by = "id") 
p2 <- plotReducedDim(uncorrected, "UMAP", colour_by = "id") 

```

```{r include = FALSE}
# now plots for batch corrected (integrated) samples 
integrated <- readRDS(here('..', 'data', 'integrated_cellassign', params$pair_ids, 'integrated_cas.rds'))

# some dim red plots, start by preparing the colors we will use 

p3 <- DimPlot(integrated, reduction = 'tsne', group.by = 'id', combine = TRUE) + 
  ggtitle(paste(params$pair_ids, 'tsne showing sample ids', sep = '/'))

p4 <- visualize_cellassign(integrated, 'tsne', 'cell_types', master_color_palette, master_cell_types) + 
  ggtitle(paste(params$pair_ids, 'tsne showing cell types', sep = '/'))

p5 <- DimPlot(integrated, reduction = 'umap', group.by = 'id', combine = TRUE) + 
  ggtitle(paste(params$pair_ids, 'umap showing sample ids', sep = '/'))
          
p6 <- visualize_cellassign(integrated, 'umap', 'cell_types', master_color_palette, master_cell_types) + 
    ggtitle(paste(params$pair_ids, 'umap showing cell types', sep = '/'))


```

### Samples **before** integration, tsne and umap  
In this plots we plot the samples together without any correction on the tsne plot to see how they are positioned. This is helpful visualizing the extent of batch effects.   
```{r echo = FALSE, message = FALSE, fig.height = 4, fig.width = 10}

gridExtra::grid.arrange(p1, p2, nrow = 1)

```

```{r include = FALSE}

# load cell assign result
sce_norm_cas_list <- lapply(id.list, function(id) readRDS(here('..', 'data', 'cellassign', id, 'sce_norm_cas.rds')))

# make seurat objects 
seur_sce_list <- lapply(sce_norm_cas_list, function(sce) as.Seurat(sce)) # make a seurat object 

# run clustering and dim reduction analysis
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) ScaleData(seurat_object))

seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindVariableFeatures(seurat_object)) 

seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunPCA(seurat_object, verbose = FALSE))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunTSNE(seurat_object, dims = 1:30, check_duplicates = FALSE))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunUMAP(seurat_object, dims = 1:30))

seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindNeighbors(seurat_object, dims = 1:10))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindClusters(seurat_object, resolution = 0.5))

# TO DO LATER 
# save the seurat objects with the clustering information

# make clustering plots
clustering_plots_tsne <- lapply(seur_sce_list, function(seurat_object) DimPlot(seurat_object, reduction = "tsne") + ggtitle(paste(unique(seurat_object$id), 'clusters', sep = ' - ')))

# make cell assign plots
# find which cell types we have, sort them alphabetically 
unique_cell_types <- lapply(seur_sce_list, function(seurat_object) as.vector(sort(unique(seurat_object$cell_type))))

# find common cell types between our lists and the master cell type list 
common_list <- lapply(unique_cell_types, function(some_list) intersect(some_list, master_cell_types))

# find the indices where we have common these cell types in the master cell type list 
idx_list <- lapply(common_list, function(some_list) match(some_list, master_cell_types))

# pick colors that correspond to the indices
colors_chosen_list <- lapply(idx_list, function(idx) master_color_palette[idx]) 

# make cell assign plots
plots_list <- list()
i <- 1 

while (i <= length(seur_sce_list)){
  
  plot <- DimPlot(object = seur_sce_list[[i]],
              dims = c(1, 2),
              reduction = 'tsne',
              group.by = 'cell_type',
              pt.size = 1.5) +
  scale_color_manual(values = colors_chosen_list[[i]]) +
  labs(title = paste(unique(seur_sce_list[[i]]$id), 'Cell Assign for scRNAseq Cells', sep = '-'))

  plots_list[[paste(i)]] <- plot # save the plot to the plots list
  
  i <- i + 1
  
}

```

### Samples **after** integration and cell assign run  
#### tsne plots  
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15}

gridExtra::grid.arrange(p3, p4, nrow = 1)

```

#### umap plots  
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15}

gridExtra::grid.arrange(p5, p6, nrow = 1)

```

Next we include some clustering and cell type plots. On the left, we show the tSNE plot with clustering information where each cluster is colored with a different color. On the right, cell assign results from the integrated sample is given.

```{r include = FALSE}

integrated <- FindNeighbors(integrated, dims = 1:10)
integrated <- FindClusters(integrated, resolution = 0.5)

clustering_plot <- DimPlot(integrated, reduction = 'tsne', group.by = 'seurat_clusters', combine = TRUE) + ggtitle(paste(params$pair_ids, 'clustering', sep = ' / '))

cell_type_plot <- visualize_cellassign(integrated, 'tsne', 'cell_types', master_color_palette, master_cell_types) + ggtitle(paste(params$pair_ids, 'cell types', sep = ' / '))


```

```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15}

grid.arrange(clustering_plot, cell_type_plot, nrow = 1)

```

### Individual samples  

#### Clustering  
Running a clustering algorithm on individual samples can help better understand possible sub cell populations in these samples. 0.5 was chosen for resolution. We show cell type and clustering results side by side. Clustering plots (tsne) are given on the left side and cell assign results are given on the right side for each separate sample given.  

```{r echo = FALSE, message = FALSE, fig.height = 6 * length(plots_list), fig.width = 15, fig.show = 'hold'}


# now we visualize the results based on the number of samples analyzed

if (length(clustering_plots_tsne) == 1) {
  grid.arrange(clustering_plots_tsne[[1]], plots_list[[1]], nrow = 1)
  
} else if (length(clustering_plots_tsne) == 2) {
  grid.arrange(clustering_plots_tsne[[1]], plots_list[[1]], 
               clustering_plots_tsne[[2]], plots_list[[2]], nrow = 2)
  
} else if (length(clustering_plots_tsne) == 3) {
  grid.arrange(clustering_plots_tsne[[1]], plots_list[[1]], 
               clustering_plots_tsne[[2]], plots_list[[2]], 
               clustering_plots_tsne[[3]], plots_list[[3]], nrow = 3)
  
} else if (length(clustering_plots_tsne) == 4) {
  grid.arrange(clustering_plots_tsne[[1]], plots_list[[1]], 
               clustering_plots_tsne[[2]], plots_list[[2]], 
               clustering_plots_tsne[[3]], plots_list[[3]],
               clustering_plots_tsne[[4]], plots_list[[4]], nrow = 4)

} else if (length(clustering_plots_tsne) == 5) {
  grid.arrange(clustering_plots_tsne[[1]], plots_list[[1]], 
               clustering_plots_tsne[[2]], plots_list[[2]], 
               clustering_plots_tsne[[3]], plots_list[[3]],
               clustering_plots_tsne[[4]], plots_list[[4]],
               clustering_plots_tsne[[5]], plots_list[[5]], nrow = 5) } # end of if loop 

```
