---
title: 'Batch Effect Correction with Scran and Seurat'
subtitle: '`r params$ids` - `r params$id_type` '
author: 'asli'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: true

params:
  ids: 'ids'
  path_to_sce_clus: 'path_to_sce_clus'
  path_to_sce_qc: 'path_to_sce_qc'
  id_type: 'id_type'
  
---

Batch effects can introduce unwanted variation to samples. It's important to normalize the batch effect before attempting to analyze samples together. Here we show some dimensionality reduction plots (tSNE and UMAP) before and after batch normalization with **scran** and **seurat** in these samples:  
`r params$ids`. 

Cells should not cluster by their source/orgin (i.e., sampleID) if the batch effect is sucessfully removed. We can compare batch correction with scran and seurat to decide which is the better correction method we would like to go with.


```{r include = FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
source(here('pipeline', 'batch_normalization', 'integration.R'))

```


## Scran batch correction

We first correct for sequencing depth by calling `multiBatchNorm()` which downscales all batches to match the coverage of the least-sequenced batch. Then, we combine sces by directly merging them and use fast mutual nearest neighbours (fastMNN) to identify and remove batch effect. Dimension reduction plots (tSNE and UMAP) are shown to compare sample clustering before and after scran batch correction. 


```{r include = FALSE, message = FALSE}

# snakemake passes the input files as one string separated by a space. we need to divide them into separate lists: 
paths <- as.list(unlist(strsplit(params$path_to_sce_clus[[1]], ' '))) 

# load the sces
sces <- lapply(paths, function(some_path) readRDS(some_path)) 

```

```{r include = FALSE}

# subset to common genes 
universe <- Reduce(intersect, lapply(sces, function(sce) rownames(sce)))
sces <- lapply(sces, function(sce) sce[universe, ])

# make the sces comparable by removing sequencing depth 
sces <- do.call(multiBatchNorm, sces)

# add a prefix to colnames of each sce 
for (i in 1:length(sces)) colnames(sces[[i]]) <- paste(unique(sces[[i]]$id), colnames(sces[[i]]), sep = ".")

# find highly variable genes  
vars <- lapply(sces, function(sce) modelGeneVar(sce))
combined.var <- do.call(combineVar, vars)
chosen.hvgs <- combined.var$bio > 0

# combine without any correction yet
combined <- do.call(correctExperiments, c(PARAM=NoCorrectParam(), sces))
combined <- runPCA(combined, subset_row = chosen.hvgs)
combined <- runTSNE(combined, dimred = "PCA")
combined <- runUMAP(combined, dimred = "PCA")

p1 <- plotTSNE(combined, colour_by = "id")
p2 <- plotUMAP(combined, colour_by = "id")

# batch correction! 
batch_removed <- fastMNN(combined, batch = combined$id, subset.row = chosen.hvgs)

batch_removed <- runTSNE(batch_removed, dimred = "corrected")
batch_removed <- runUMAP(batch_removed, dimred = "corrected")

p3 <- plotTSNE(batch_removed, colour_by = "batch")
p4 <- plotUMAP(batch_removed, colour_by = "batch")

```

#### **Before** batch correction  
```{r echo = FALSE, message = FALSE, fig.height = 4, fig.width = 8}

gridExtra::grid.arrange(p1, p2, nrow = 1)

```

#### **After** batch correction  
```{r echo = FALSE, message = FALSE, fig.height = 4, fig.width = 8}

gridExtra::grid.arrange(p3, p4, nrow = 1)

```



## Seurat integration and batch correction

We can integrate samples using Seurat integration. Seurat first finds integration anchors which are cells that have similar biological state across samples. Then, individual samples are integrated into one integrated sample on which dimension reduction and clustering are ran. Best number of clusters is decided by seurat.


```{r include = FALSE}

# snakemake passes the input files as one string separated by a space. we need to divide them into separate lists: 
path_to_sce_qc <- as.list(unlist(strsplit(params$path_to_sce_qc[[1]], ' '))) 
path_to_sce_clus <- as.list(unlist(strsplit(params$path_to_sce_clus[[1]], ' '))) 

# perform seurat integration and batch correction
out <- integrate(path_to_sce_qc = path_to_sce_qc, 
                 path_to_sce_clus = path_to_sce_clus, 
                 ids_integration = params$ids)

# we start with the plots for uncorrected samples  
uncorrected <- out$uncorrected

# some dim red plots 
p1 <- plotReducedDim(uncorrected, "TSNE", colour_by = "id") 
p2 <- plotReducedDim(uncorrected, "UMAP", colour_by = "id") 


# now plots for batch corrected (integrated) samples 
integrated <- out$corrected

# some dim red plots 
p3 <- DimPlot(integrated, reduction = 'tsne', group.by = 'id', combine = FALSE)
p4 <- DimPlot(integrated, reduction = 'umap', group.by = 'id', combine = FALSE)

p3 <- ggplotify::as.grob(p3[[1]])
p4 <- ggplotify::as.grob(p4[[1]])

```

#### **Before** seurat batch correction
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 10}

gridExtra::grid.arrange(p1, p2, nrow = 1)

```

#### **After** seurat batch correction  
```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 10}

gridExtra::grid.arrange(p3, p4, nrow = 1)

```



