---
title: "Cell assign results"
subtitle: '`r params$ids`'
author: "Asli Munzur"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  ids: 'ids'
  sce_clus: 'sce_clus'
  cell_type_csv: 'cell_type_csv'
  dm_path: 'dm_path'
  
---

```{r include=FALSE}

library(here)
library(plotly)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
source(here('pipeline', 'cellassign', 'scripts', 'create_marker_matrix_GC.R'))
source(here('pipeline', 'cellassign', 'scripts', 'assign_cell_type.R'))


```

```{r echo = FALSE, message = FALSE, warning = FALSE}

sce_clus <- readRDS(params$sce_clus) # load sce 
cell_type_csv <- read_csv(params$cell_type_csv) # load cell type info 
sce_clus$cell_type <- cell_type_csv


# in case we need to redo clustering step 
# set.seed(12)
# if (length(sce_clus$cluster) == 0 ) {
#   
#   g <- buildSNNGraph(sce_clus, k = 25, use.dimred = 'PCA')
#   clust <- igraph::cluster_walktrap(g)$membership
#   sce_clus$cluster <- factor(clust)
#   
# }

table(sce_clus$cluster) %>% 
  knitr::kable(caption = 'Table 1: number of cells in each cluster', 
               col.names = c('cluster', 'number of cells')) %>% 
  kable_styling(full_width = F)


```

```{r test-rgl, webgl=TRUE, message = FALSE, warning = FALSE}

# if dim reduction hasnt been computed, do it now: 
# if (length(sce_clus@int_colData@listData[["reducedDims"]]$TSNE) == 0 ) {
#   
#   sce_clus <- runPCA(sce_clus, exprs_values = "logcounts", ncomponents = 50)
#   sce_clus <- runTSNE(sce_clus, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, perplexity = 20)
#   sce_clus <- runUMAP(sce_clus, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, min_dist = 0.5, n_neighbors = 15, metric = "euclidean") 
#   
# } 

# get tsne coordinates 
tsne_1 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$TSNE[, 1])
tsne_2 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$TSNE[, 2])
tsne_3 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$TSNE[, 3])

# get pca coordinates 
pca_1 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$PCA[, 1])
pca_2 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$PCA[, 2])
pca_3 <- as.numeric(sce_clus@int_colData@listData[["reducedDims"]]$PCA[, 3])

df <- data.frame(tsne_1, tsne_2, tsne_3)
df$cell_type <- as.factor(sce_clus$cell_type$cell_type)
df$cluster <- as.factor(sce_clus$cluster)

colors_list = c(brewer.pal(n = 8, name = "Set2"), brewer.pal(n = 8, name = "Paired"))


if (length(unique(sce_clus$cell_type)) > 8) {
  
  # 2d tsne plot showing cell types 
  cell_type_plot <- plot_ly(data = df,
          x = ~tsne_1, y = ~tsne_2,
          opacity = 1,
          color = ~cell_type,
          type = "scatter",
          mode = "markers",
          marker = list(size = 4), 
          colors = colors_list) %>%  
          layout(title = 'cell types in the sample shown in t-SNE plot', 
                 legend= list(itemsizing='constant'))

} else {
  
    # 2d tsne plot showing cell types 
  cell_type_plot <- plot_ly(data = df,
          x = ~tsne_1, y = ~tsne_2,
          opacity = 1,
          color = ~cell_type,
          type = "scatter",
          mode = "markers",
          marker = list(size = 4)) %>%  
          layout(title = 'cell types in the sample shown in t-SNE plot', 
                 legend= list(itemsizing='constant'))

}


# 2d plot showing clusters
cluster_plot <- plot_ly(data = df,
        x = ~tsne_1, y = ~tsne_2,
        opacity = 1,
        color = ~cluster,
        type = "scatter",
        mode = "markers",
        marker = list(size = 4), 
        colors = "Set1") %>%  
        layout(title = 'clusters in the sample shown in t-SNE plot', 
               legend= list(itemsizing='constant'))

```

### Cell assign results  
Below the cell assign results are shown in t-SNE plots. Also, another t-SNE plot is added to show the clustering results since comparing cell types to different clusters that came up might provide additional insights. The t-SNE plots are interactive, when you hover over the plot, you'll notice that a panel comes up in the upper right corner of the plots where a variety of options are given, like zoom in, zoom out etc.  

One of the useful tools is to select the area you wish to zoom in, then the plots will automatically take you there.  

```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 7, warning = FALSE}

cell_type_plot
cluster_plot

```

### Ways to interact with the 3d plots below  
Same information in the plots above is also presented in interactive 3d plots below. Some of the things you can do with the interactive plots below are:  

 - Zoom in and zoom out using the mouse wheel  
 - Click on a region in the plot and drag to see different angles  
 - Double click on a cell type (or cluster number) to hide everyhing else in the plots and show that cell type only. To go back to how the plot was before, double click on the same place in the legend. If you wish to highlight more than one cell type, double click to one of them in the legend, then single click the other one. You can also do this in the 2d plots above as well. 
 - If you bring your mouse pointer over the 3d plots, you'll notice that a faint panel on the right upper corner of the plot appears. That panel gives more options to interact with the plot. To reset everything in the plot, click "reset camera to last save".  


```{r, message = FALSE, warning = FALSE}

if (length(unique(sce_clus$cell_type)) > 8) {
  
  p1 <- plot_ly(data = df,
        x = ~tsne_1, y = ~tsne_2, z = ~tsne_3,
        opacity = 1,
        color = ~cell_type,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3), 
        colors = colors_list) %>% 
  layout(title = '3d t-SNE plot showing cell types', 
         legend= list(itemsizing='constant'))


} else {
  
  p1 <- plot_ly(data = df,
          x = ~tsne_1, y = ~tsne_2, z = ~tsne_3,
          opacity = 1,
          color = ~cell_type,
          type = "scatter3d",
          mode = "markers",
          marker = list(size = 3)) %>% 
    layout(title = '3d t-SNE plot showing cell types', 
           legend= list(itemsizing='constant'))

}

p2 <- plot_ly(data = df,
        x = ~tsne_1, y = ~tsne_2, z = ~tsne_3,
        opacity = 1,
        color = ~cluster,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3), 
        colors = "Set1") %>% 
  layout(title = '3d t-SNE plot showing clusters', 
         legend= list(itemsizing='constant'))


```

```{r message = FALSE, warning = FALSE}

p1
p2

```

#### Diffusion maps  
Diffusion maps are a useful technique to pick up branching events in the sample where cells differentiate into different cell types. Below cell type information is shown in both 2 and 3 dimensional diffusion maps. You can interact with these plots same way as above. 

```{r include = FALSE}

# load the dm from params 
dm <- readRDS(params$dm_path)

# make df with the diffusion map coodinates, we will use this later to make the dm plot 
tmp <- data.frame(DC1 = eigenvectors(dm)[, 1],
                  DC2 = eigenvectors(dm)[, 2],
                  DC3 = eigenvectors(dm)[, 3],
                  cell_type = sce_clus$cell_type$cell_type) # this will help color based on cluster 

if (length(unique(sce_clus$cell_type)) > 8) {
  
    # make the 2D plot 
  dm_2d <- plot_ly(data = tmp,
          x = ~DC1, y = ~DC2,
          opacity = 1,
          color = ~cell_type,
          type = "scatter",
          mode = "markers",
          marker = list(size = 5), 
          colors = colors_list) %>% 
    layout(title = '2D diffusion map showing cell types', 
           legend= list(itemsizing='constant'))

} else {
  
  dm_2d <- plot_ly(data = tmp,
          x = ~DC1, y = ~DC2,
          opacity = 1,
          color = ~cell_type,
          type = "scatter",
          mode = "markers",
          marker = list(size = 5)) %>% 
    layout(title = '2D diffusion map showing cell types', 
           legend= list(itemsizing='constant'))

}


if (length(unique(sce_clus$cell_type)) > 8) {
  
  # make the 3d plot 
  dm_3d <- plot_ly(data = tmp,
          x = ~DC1, y = ~DC2, z = ~DC3,
          opacity = 1,
          color = ~cell_type,
          type = "scatter3d",
          mode = "markers",
          marker = list(size = 3), 
          colors = colors_list)  %>% 
    layout(title = '3D diffusion map showing cell types', 
           legend= list(itemsizing='constant'))

} else {
  
  # make the 3d plot 
  dm_3d <- plot_ly(data = tmp,
          x = ~DC1, y = ~DC2, z = ~DC3,
          opacity = 1,
          color = ~cell_type,
          type = "scatter3d",
          mode = "markers",
          marker = list(size = 3))  %>% 
    layout(title = '3D diffusion map showing cell types', 
           legend= list(itemsizing='constant'))

}

```

```{r warning = FALSE, message = FALSE}

dm_2d
dm_3d

```







