---
title: Cell assign results 
subtitle: '`r params$ids_integration` - `r params$id_type`'
author: "aslı"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4

params:
  ids_integration: 'ids_integration' # write here the ids you want to integrate, separate by '-'
  id_type: 'id_type' # here you can write a quick summary of what these ids_integration are, like tissue types etc. 
  cell_types: 'cell_types'  
  sce_clus: 'sce_clus'
  cell_type_csv: 'cell_type_csv'
  seurat_corrected: 'seurat_corrected'
  scran_corrected: 'scran_corrected'
  integrate: 'integrate'

---

In this report we show the results of cell assign runs on individual samples as well as on integrated and batch corrected samples. We include some dim reduction plots with cell types and clustering information. We first show results of cell assign run on individual samples, then we give the plots for integrated samples.   


```{r include = FALSE, message = FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))
source(here('pipeline', 'cellassign', 'scripts', 'run_cellassign.R'))
source(here('pipeline', 'cellassign', 'scripts', 'create_marker_matrix.R'))
set.seed(6666)


```


```{r include = FALSE}

# get separate ids 
id.list <- strsplit(params$ids_integration, "-")[[1]] # split by "-" SO ID must NOT CONTAIN "-"
id.list <- as.list(c(unlist(list(id.list))))

# snakemake passes the input files as one string separated by a space. we need to divide them into separate lists: 
sce_clus <- as.list(unlist(strsplit(params$sce_clus[[1]], ' '))) 
cell_type_csv <-  as.list(unlist(strsplit(params$cell_type_csv[[1]], ' '))) 
cell_types <- strsplit(as.character(params$cell_types), " ")[[1]]
```

```{r include = FALSE}
# load cell_type_csv if object already exists; or run cellassign
sce_cas <- list()
cell_type <- list()
for(i in 1:length(id.list)){
  if(file.exists(cell_type_csv[[i]])==TRUE){
    sce_cas[[i]] <- readRDS(sce_clus[[i]])
    cell_type[[i]] <- read_csv(cell_type_csv[[i]])
    # add cell type information to sce_cas:
    sce_cas[[i]]$cell_type <- cell_type[[i]]$cell_type
  } else {
    # run cell assign and get sce with cell type information
    sce_cas[[i]] <- run_cellassign(sce_clus[[i]], cell_types, cell_type_csv[[i]])
  }
}
```

```{r include = FALSE}
# chunk option: if one sample, returns FALSE
eval_text <- FALSE

if(length(id.list) > 1 & params$integrate =="yes"){
  # load seurat batch corrected object
  integrated <- readRDS(params$seurat_corrected)

  # load scran object
  scran <- readRDS(params$scran_corrected)
  
  # chunk option: if two samples, returns TRUE
  eval_text <- TRUE
} 
```

```{r include = FALSE}
# get marker matrix and sort 
marker_mat <- create_marker_mat(cell_types)
det_cell_type <- colnames(marker_mat)
master_cell_types <- colnames(marker_mat) %>% str_sort()
```


**Cell assign is able to detect the following cell types:**

`r det_cell_type`



## Individual samples

### Clustering
Number of cells in each cluster is summarized in the tables below.

```{r, results='asis'}

t <- list()
for(i in 1:length(sce_cas)){
  t[[i]] <- table(sce_cas[[i]]$cluster) %>%
    knitr::kable(caption = paste(id.list[[i]], ': Number of cells in each cluster'),
                 col.names = c('cluster', 'number of cells')) %>%
    kable_styling(full_width = F)

  print(t[[i]])
}

```


### Dim reduction plots
Running a clustering algorithm on individual samples can help better understand possible sub cell populations in these samples. 0.5 was chosen for resolution. We show cell type and clustering results side by side. Clustering plots are given on the left side and cell assign results are given on the right side for each sample.

```{r include = FALSE}

# make seurat objects
seur_sce_list <- lapply(sce_cas, function(sce) as.Seurat(sce))

# run clustering and dim reduction analysis
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) ScaleData(seurat_object))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindVariableFeatures(seurat_object))

seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunPCA(seurat_object, verbose = FALSE))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunTSNE(seurat_object, dims = 1:30, check_duplicates = FALSE))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) RunUMAP(seurat_object, dims = 1:30))

seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindNeighbors(seurat_object, dims = 1:10))
seur_sce_list <- lapply(seur_sce_list, function(seurat_object) FindClusters(seurat_object, resolution = 0.5))


# make clustering plots
clustering_plots_tsne <- lapply(seur_sce_list, function(seurat_object) DimPlot(seurat_object, reduction = "tsne") +
                                  ggtitle(paste(unique(seurat_object$id), 'tSNE: clustering', sep = ' - ')))

clustering_plots_umap <- lapply(seur_sce_list, function(seurat_object) DimPlot(seurat_object, reduction = "umap") +
                                  ggtitle(paste(unique(seurat_object$id), 'UMAP: clustering', sep = ' - ')))

# make cell assign plots
# find which cell types we have, sort them alphabetically
unique_cell_types <- lapply(seur_sce_list, function(seurat_object) as.vector(sort(unique(seurat_object$cell_type))))

# find common cell types between our lists and the master cell type list
common_list <- lapply(unique_cell_types, function(some_list) intersect(some_list, master_cell_types))

# find the indices where we have common these cell types in the master cell type list
idx_list <- lapply(common_list, function(some_list) match(some_list, master_cell_types))

# pick colors that correspond to the indices
colors_chosen_list <- lapply(idx_list, function(idx) master_color_palette[idx])

# make cell assign plots
plots_list_tsne <- list()
plots_list_umap <- list()
for(i in 1:length(seur_sce_list)){
  plots_list_tsne[[i]] <- DimPlot(object = seur_sce_list[[i]],
              dims = c(1, 2),
              reduction = 'tsne',
              group.by = 'cell_type',
              pt.size = 1.5) +
  #scale_color_manual(values = colors_chosen_list[[i]]) +
  labs(title = paste(unique(seur_sce_list[[i]]$id), 'tSNE: Cell Assign for scRNAseq Cells', sep = ' - '))

    plots_list_umap[[i]] <- DimPlot(object = seur_sce_list[[i]],
              dims = c(1, 2),
              reduction = 'umap',
              group.by = 'cell_type',
              pt.size = 1.5) +
  #scale_color_manual(values = colors_chosen_list[[i]]) +
  labs(title = paste(unique(seur_sce_list[[i]]$id), 'UMAP: Cell Assign for scRNAseq Cells', sep = ' - '))
}

```

#### tSNE plots
```{r echo = FALSE, fig.width = 15, fig.show = 'hold'}

# now we visualize the results based on the number of samples analyzed

for (i in 1:length(clustering_plots_tsne)) {
  grid.arrange(clustering_plots_tsne[[i]], plots_list_tsne[[i]], nrow = 1)
}

```


#### UMAP plots
```{r echo = FALSE, fig.width = 15, fig.show = 'hold'}

for (i in 1:length(clustering_plots_tsne)) {
  grid.arrange(clustering_plots_umap[[i]], plots_list_umap[[i]], nrow = 1)
}

```



## Integrated sample

Next, we include some clustering and cell type plots. On the left, we show dim reduction plot with clustering information where each cluster is colored with a different color. On the right, cell assign results from the integrated sample is given.

### Seurat integrated sample
Below are tsne and umap plots showing cell assignment results for sample integrated and corrected with seurat.

```{r include = FALSE, warning=FALSE, eval=eval_text}
# subset to common genes across a group of sces
sce_cas <- intersect_all(sce_cas)

# append the id name in front of the barcodes
for (i in 1:length(sce_cas)) {
  colnames(sce_cas[[i]]) <- paste(id.list[[i]], colnames(sce_cas[[i]]), sep = '.')
}

# add cell types to the integrated object
celltype <- lapply(sce_cas, function(sce) sce$cell_type)
celltype_barcodes <- lapply(sce_cas, function(sce) colnames(sce))
celltype_df <- data.frame(cell_type = unlist(celltype))
rownames(celltype_df) <- unlist(celltype_barcodes)
integrated <- AddMetaData(integrated, celltype_df, col.name = 'cell_type')

# dim reduction plots
clustering_plot_tsne <- DimPlot(integrated, reduction = 'tsne', group.by = 'seurat_clusters', combine = TRUE) +
  ggtitle(paste(params$ids_integration, 'tSNE - clustering', sep = ': '))

cell_type_plot_tsne <- visualize_cellassign(integrated, 'tsne', 'cell_type', master_color_palette, master_cell_types) +
  ggtitle(paste(params$ids_integration, 'tSNE - cell types', sep = ': '))


clustering_plot_umap <- DimPlot(integrated, reduction = 'umap', group.by = 'seurat_clusters', combine = TRUE) +
  ggtitle(paste(params$ids_integration, 'UMAP - clustering', sep = ': '))

cell_type_plot_umap <- visualize_cellassign(integrated, 'umap', 'cell_type', master_color_palette, master_cell_types) +
  ggtitle(paste(params$ids_integration, 'UMAP - cell types', sep = ': '))

```

#### tSNE plots
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15, eval=eval_text}

grid.arrange(clustering_plot_tsne, cell_type_plot_tsne, nrow = 1)

```


#### UMAP plots
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15, eval=eval_text}

grid.arrange(clustering_plot_umap, cell_type_plot_umap, nrow = 1)

```


### Scran combined sample
Below are tsne and umap plots showing clustering and cell assignment results for sample integrated and corrected with scran

```{r include = FALSE, warning=FALSE, eval=eval_text}

# add cell type information to combined object
scran$cell_type <- celltype_df$cell_type

# set color palette
unique_cell_types <- as.vector(sort(unique(scran$cell_type)))
common_list <- intersect(unique_cell_types, master_cell_types)
idx_list <- match(common_list, master_cell_types)
colors_chosen <- master_color_palette[idx_list]

# dim reduction plots
clustering_plot_tsne <- plotTSNE(scran, colour_by="cluster") + ggtitle(paste(params$ids_integration, 'TSNE - clustering', sep = ': '))
cell_type_plot_tsne <- plotTSNE(scran, colour_by="cell_type") + ggtitle(paste(params$ids_integration, 'TSNE - cell type', sep = ': ')) +
    scale_fill_manual(values = colors_chosen)

clustering_plot_umap <- plotUMAP(scran, colour_by="cluster") + ggtitle(paste(params$ids_integration, 'UMAP - clustering', sep = ': '))
cell_type_plot_umap <- plotUMAP(scran, colour_by="cell_type") + ggtitle(paste(params$ids_integration, 'UMAP - cell type', sep = ': '))  +
    scale_fill_manual(values = colors_chosen)

```

#### tSNE plots
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15, eval=eval_text}

grid.arrange(clustering_plot_tsne, cell_type_plot_tsne, nrow = 1)

```

#### UMAP plots
```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 15, eval=eval_text}

grid.arrange(clustering_plot_umap, cell_type_plot_umap, nrow = 1)

```






