---
title: "Dimensionality reduction plots "
subtitle: '`r params$sample`'
author: "Asli Munzur"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide

params:
  path_to_sce_norm: 'path_to_sce_norm'
  sce_red: 'sce_red'
  seed1: 'seed1'
  seed2: 'seed2'
  HVG: 'HVG'
  top_HVG: 'top_HVG'
  top_PCs: 'top_PCs'
  
---


```{r include = FALSE}

library(here)
source(here('pipeline', 'sourceFiles', 'utilities.R'))

```
  
  
  
```{r}

# load sce
sce_norm <- readRDS(params$path_to_sce_norm)

# compute the number of PCs to use 
if (params$top_PCs == "computed") { # pipeline decides on the number of PCs to use 
  
  set.seed(params$seed1)
  dec.sce <- modelGeneVarByPoisson(sce_norm) # model gene variation, decompose it into technical and biological components
  
  # choose the HVG
  fraction_HGV <- as.numeric(params$top_HVG) / 100
  top.sce <- getTopHVGs(dec.sce, prop = fraction_HGV)
  
  denoised.sce <- denoisePCA(sce_norm, technical = dec.sce)
  PCs <- ncol(reducedDim(denoised.sce))
  
} else { # user decided how many PCs to use 
  
  top_PCs = as.numeric(params$top_PCs)
  PCs = top_PCs # number of PCs to use is given by the user in the top_PCs parameter
}

# deal with HVG
if (params$HVG == "yes") { # user wants to subset the sce to HVG 
  
  # pick the HVG
  gene_var <- modelGeneVar(sce_norm)
  
  fraction_HGV <- as.numeric(params$top_HVG) / 100
  chosen <- getTopHVGs(gene_var, prop = fraction_HGV)
  
  # subset to HVG
  sce_qc_hvg <- sce_norm[chosen,]
  
  # add the original count matrix with non HVG to our new sce
  altExp(sce_qc_hvg, "original") <- sce_norm
  sce_norm <- sce_qc_hvg
  
  set.seed(params$seed2)
  sce_norm <- runPCA(sce_norm, exprs_values = "logcounts", ncomponents = PCs)
  
} else { # compute PCA (and other dim reduction methods) without subsetting to top HVG
  
  set.seed(params$seed2)
  sce_norm <- runPCA(sce_norm, exprs_values = "logcounts", ncomponents = PCs)
  
}

# TSNE and UMAP 
set.seed(params$seed2)
sce_norm <- runTSNE(sce_norm, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, perplexity = 20)
sce_norm <- runUMAP(sce_norm, dimred = "PCA", exprs_values = "logcounts", ncomponents = 3, min_dist = 0.5, n_neighbors = 15, metric = "euclidean")

# save the output 
saveRDS(sce_norm, file = params$sce_red)


```

```{r include = FALSE}

# some seurat plots here to decide how many PCs to include 
# convert sce to seurat object
sobject <- as.Seurat(sce_norm, counts = "counts", data = "logcounts")
Idents(sobject) <- paste(unique(sce_norm$id))

sobject <- ScaleData(sobject)
sobject <- FindVariableFeatures(sobject)

sobject@reductions$pca <- sobject@reductions$PCA
sobject@reductions$tsne <- sobject@reductions$TSNE
sobject@reductions$umap <- sobject@reductions$UMAP

sobject <- RunPCA(sobject, features = VariableFeatures(object = sobject))

```

#### Visualize top genes associated with reduction components, in top 6 PCs
```{r, fig.width = 14, fig.height = 10}

VizDimLoadings(sobject, dims = 1:6, reduction = "pca", ncol = 3)

```

### Deciding how many PCs to include in further analysis  
In the following heatmaps top genes contributing to each PC are visualized. Both cells and genes are ordered by their PC scores, and the 15 genes with the highest and 15 genes with the lowest PC scores are displayed for each PC. Note that the cells are given in the columns and genes are show in the rows. Here the goal is to pick the PCs that can divide the cell population into distinct groups. If all of the cells are the same color (pink), then that PC isn't as significant. 

```{r fig.width = 16, fig.height = 25}

DimHeatmap(sobject, dims = 1:15, cells = 500, balanced = TRUE)

```

### Finding significant PCs
‘Significant’ PCs will show a strong enrichment of genes with low p-values (solid curve above the dashed line) in the plot below.  

```{r fig.width = 10, fig.height = 7, warning = FALSE, message=FALSE}

sobject <- JackStraw(object = sobject, reduction = "pca", dims = 20, num.replicate = 100,  prop.freq = 0.1, verbose = FALSE)

sobject <- ScoreJackStraw(object = sobject, dims = 1:20, reduction = "pca")
JackStrawPlot(object = sobject, dims = 1:20, reduction = "pca")


```

To further validate the results of the plot above, we can also check out contribution of each PC to total standard deviation and find the elbow point.  
```{r fig.width = 10, fig.height = 4}

ElbowPlot(object = sobject)

```














